---
title: "tables"
format: html
editor: visual
---

```{r}
library(readr)
library(tidyverse)
library(lubridate)
library(splines)
library(broom)
library(usmap)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(officer)
library(flextable)
library(dplyr)
library(conflicted)
conflicts_prefer(dplyr::select)

```

## Join

```{r}
N172010 <- read_csv("Data/N172010-1.csv") %>%
  filter(`Year Code` < 2018)
N172018 <- read_csv("Data/N172018-1.csv")
clean_year <- function(df) {
  df %>%
    mutate(
      Year_raw = as.character(Year),
      # pull the first 4-digit year (e.g., "2024 (provisional)" -> 2024)
      Year = readr::parse_number(Year_raw),
      # keep any extra text as a note (optional)
      Year_note = str_squish(str_remove(Year_raw, "^\\d{4}"))
    )
}

N17 <- bind_rows(
  clean_year(N172010 ),
  clean_year(N172018 )
) %>%
  select(-Year_raw)

glimpse(N17)
```

```{r}
A402010 <- read_csv("Data/A402010-1.csv")%>%
  filter(`Year Code` < 2018)
A402018 <- read_csv("Data/A402018-1.csv")
A40 <- bind_rows(
  clean_year(A402010 ),
  clean_year(A402018 )
) %>%
  select(-Year_raw)
```

```{r}
N17A402010 <- read_csv("Data/N17A402010-1.csv")%>%
  filter(`Year Code` < 2018)
N17A402018 <- read_csv("Data/N17A402018-1.csv")

N17A40 <- bind_rows(
  clean_year(N17A402010 ),
  clean_year(N17A402018 )
) %>%
  select(-Year_raw)
N17A40_RF <- N17A40 %>%
  dplyr::filter(`MCD - ICD Sub-Chapter` == "Renal failure")
```

## Year table

```{r}
std2000_allages <- tibble(
  age_code = c(
    "1","1-4",
    "5-9","10-14",
    "15-19","20-24",
    "25-29","30-34",
    "35-39","40-44",
    "45-49","50-54",
    "55-59","60-64",
    "65-69","70-74",
    "75-79","80-84",
    "85+"
  ),
  std_pop = c(
    3794901, 15191619,
    19919840, 20056779,
    19819518, 18257225,
    17722067, 19511370,
    22179956, 22479229,
    19805793, 17224359,
    13307234, 10654272,
    9409940, 8725574,
    7414559, 4900234,
    2678567 + 1580606  # 85-89 + 90+
  )
) %>%
  mutate(weight = std_pop / sum(std_pop))
```

```{r}
# ============================================================
# make_year_table_with_aamr() — ROBUST AAMR (all ages)
#
# Fixes the 2018+ “AAMR too low” problem by:
#   1) Computing AAMR from *non-stratified* ("All") rows ONLY
#      (Sex = All, Race/Single Race 6 = All) to avoid provisional
#      suppression bias when summing stratified cells.
#   2) Using the official 2000 U.S. standard population in the
#      *published age groups* (mostly 10-year bins), by collapsing
#      WONDER 5-year ages into those bins (avoids incorrect /2 splits).
#
# You can still compute sex/race counts from stratified rows,
# but DO NOT use stratified rows for AAMR in 2018+.
# ============================================================

make_year_table_with_aamr <- function(df,
                                      year_min = 2010,
                                      year_max = 2025,
                                      subchapter_filter = NULL,
                                      add_total_row = TRUE,
                                      verbose = TRUE) {

  suppressPackageStartupMessages({
    library(dplyr)
    library(tidyr)
    library(tibble)
    library(readr)
    library(stringr)
  })

  # ============================================================
  # 2000 U.S. standard population (published groups; sums to 1,000,000)
  # Source: NCHS/SEER standard million (commonly used by CDC/WONDER)
  # Groups here align to 10-year bins (plus <1, 1–4, 85+).
  # ============================================================
  std2000_allages <- tibble(
    std_age = c("1", "1-4",
                "5-14","15-24","25-34","35-44","45-54","55-64","65-74","75-84","85+"),
    std_million = c(13818, 55317,
                    145565, 138646, 135573, 162613, 134834,  87247,  66037,  44842, 15508)
  ) %>%
    mutate(weight = std_million / sum(std_million))

  # ============================================================
  # Helpers
  # ============================================================
  parse_pop_safe <- function(x) {
    x <- na_if(as.character(x), "Not Applicable")
    readr::parse_number(x, na = c("", "NA", "Not Applicable"))
  }

  # Recode 5-year age codes to the *standard* age groups above
  # Works whether WONDER uses:
  #   - "1" and "1-4"
  #   - 5-year bins like "5-9", "10-14", ...
  #   - older bins "85-89", "90-94", "95-99", "100+"
  recode_to_std_age <- function(age_code_raw) {
    age_code_raw <- as.character(age_code_raw)

    dplyr::case_when(
      age_code_raw %in% c("1", "<1", "Under 1 year") ~ "1",
      age_code_raw %in% c("1-4") ~ "1-4",

      age_code_raw %in% c("5-9","10-14") ~ "5-14",
      age_code_raw %in% c("15-19","20-24") ~ "15-24",
      age_code_raw %in% c("25-29","30-34") ~ "25-34",
      age_code_raw %in% c("35-39","40-44") ~ "35-44",
      age_code_raw %in% c("45-49","50-54") ~ "45-54",
      age_code_raw %in% c("55-59","60-64") ~ "55-64",
      age_code_raw %in% c("65-69","70-74") ~ "65-74",
      age_code_raw %in% c("75-79","80-84") ~ "75-84",

      age_code_raw %in% c("85+","85-89","90-94","95-99","100+") ~ "85+",
      TRUE ~ NA_character_
    )
  }

  # Identify which variables partition population in the extract
  # (Used only for “dedupe pop by stratum” logic.)
  get_part_vars <- function(dat) {
    pv <- intersect(c("Sex", "Single Race 6", "Race"), names(dat))
    if (all(c("Race", "Single Race 6") %in% pv)) pv <- setdiff(pv, "Race")
    pv
  }

  # Safely detect "All" level(s) if present
  has_all_level <- function(x) {
    if (is.null(x)) return(FALSE)
    any(as.character(x) %in% c("All", "All persons", "All races"))
  }

  # Filter to "All" strata rows for AAMR computation (if columns exist)
  filter_to_all_strata <- function(dat) {
    out <- dat
    if ("Sex" %in% names(out)) out <- out %>% filter(Sex %in% c("All", NA))
    if ("Single Race 6" %in% names(out)) out <- out %>% filter(`Single Race 6` %in% c("All", NA))
    if ("Race" %in% names(out)) out <- out %>% filter(Race %in% c("All", NA))
    out
  }

  # Core: build Year × std_age totals with correct population handling
  # If pv is empty -> just Year × std_age.
  # If pv exists -> Year × std_age × pv, dedupe pop by max(), then sum across pv.
  year_stdage_totals <- function(dat) {
    pv <- get_part_vars(dat)

    if (length(pv) == 0) {
      dat %>%
        group_by(Year, std_age) %>%
        summarise(
          deaths = sum(Deaths, na.rm = TRUE),
          pop    = suppressWarnings(max(Population, na.rm = TRUE)),
          .groups = "drop"
        ) %>%
        mutate(pop = ifelse(is.infinite(pop), NA_real_, pop))
    } else {
      dat %>%
        group_by(across(all_of(c("Year", "std_age", pv)))) %>%
        summarise(
          deaths = sum(Deaths, na.rm = TRUE),
          pop    = suppressWarnings(max(Population, na.rm = TRUE)),
          .groups = "drop"
        ) %>%
        mutate(pop = ifelse(is.infinite(pop), NA_real_, pop)) %>%
        group_by(Year, std_age) %>%
        summarise(
          deaths = sum(deaths, na.rm = TRUE),
          pop    = sum(pop,    na.rm = TRUE),
          .groups = "drop"
        )
    }
  }

  # ============================================================
  # Base filter + types
  # ============================================================
  base <- df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      Year       = suppressWarnings(as.integer(`Year Code`)),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population)
    ) %>%
    filter(Year >= year_min, Year <= year_max)

  if (!is.null(subchapter_filter)) {
    base <- base %>% filter(`MCD - ICD Sub-Chapter` %in% subchapter_filter)
  }

  # ============================================================
  # Race harmonization + std_age recode
  # ============================================================
  base <- base %>%
    mutate(
      race_source = coalesce(`Single Race 6`, Race),
      race_unified = case_when(
        race_source == "American Indian or Alaska Native" ~ "American Indian or Alaska Native",
        race_source %in% c("Asian", "Asian or Pacific Islander") ~ "Asian",
        race_source == "Native Hawaiian or Other Pacific Islander" ~ "Native Hawaiian or Other Pacific Islander",
        race_source == "Black or African American" ~ "Black or African American",
        race_source == "White" ~ "White",
        race_source == "More than one race" ~ "More than one race",
        TRUE ~ NA_character_
      ),
      age_code_raw = `Five-Year Age Groups Code`,
      std_age      = recode_to_std_age(age_code_raw)
    )

  # Keep only ages that can be standardized
  base_std <- base %>%
    filter(std_age %in% std2000_allages$std_age)

  # ============================================================
  # ✅ AAMR computation dataset:
  # Prefer "All" strata rows to avoid 2018+ suppression bias
  # ============================================================
  base_for_aamr <- base_std

  # If extract appears stratified and has All-levels, use All-only rows for AAMR
  can_filter_all <- TRUE
  if ("Sex" %in% names(base_for_aamr)) can_filter_all <- can_filter_all && has_all_level(base_for_aamr$Sex)
  if ("Single Race 6" %in% names(base_for_aamr)) can_filter_all <- can_filter_all && has_all_level(base_for_aamr$`Single Race 6`)
  if ("Race" %in% names(base_for_aamr)) can_filter_all <- can_filter_all && has_all_level(base_for_aamr$Race)

  if (can_filter_all) {
    base_for_aamr <- filter_to_all_strata(base_for_aamr)
    if (verbose) message("AAMR: using All-strata rows only (avoids 2018+ suppression bias).")
  } else {
    if (verbose) {
      message(
        "AAMR: could not confirm presence of 'All' rows for every stratifier.\n",
        "If your extract is stratified (Sex/Race), AAMR for 2018+ can be biased downward due to suppressed cells.\n",
        "Best fix: re-pull WONDER for Year×Age only with Sex=All and Race=All, then rerun."
      )
    }
  }

  # Year × std_age totals (dedup pop properly)
  ya <- year_stdage_totals(base_for_aamr)

  # AAMR by year (direct standardization)
  aamr_by_year <- ya %>%
    left_join(std2000_allages, by = "std_age") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    group_by(Year) %>%
    summarise(
      `AAMR (/100,000), age-adjusted (all ages)` = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    )

  # Yearly population consistent with AAMR denominator
  pop_by_year <- ya %>%
    group_by(Year) %>%
    summarise(Population = sum(pop, na.rm = TRUE), .groups = "drop")

  # ============================================================
  # Counts table (can use stratified base; suppression affects 2018+ counts too)
  # ============================================================
  out <- base %>%
    group_by(Year) %>%
    summarise(
      `Overall (n)` = sum(Deaths, na.rm = TRUE),
      `Women (n)`   = sum(Deaths[Sex == "Female"], na.rm = TRUE),
      `Men (n)`     = sum(Deaths[Sex == "Male"],   na.rm = TRUE),

      `American Indian or Alaska Native (n)` =
        sum(Deaths[race_unified == "American Indian or Alaska Native"], na.rm = TRUE),
      `Asian (n)` =
        sum(Deaths[race_unified == "Asian"], na.rm = TRUE),
      `Native Hawaiian or Other Pacific Islander (n)` =
        sum(Deaths[race_unified == "Native Hawaiian or Other Pacific Islander"], na.rm = TRUE),
      `Black or African American (n)` =
        sum(Deaths[race_unified == "Black or African American"], na.rm = TRUE),
      `White (n)` =
        sum(Deaths[race_unified == "White"], na.rm = TRUE),
      `More than one race (n)` =
        sum(Deaths[race_unified == "More than one race"], na.rm = TRUE),

      .groups = "drop"
    ) %>%
    left_join(pop_by_year, by = "Year") %>%
    mutate(
      `Crude rate (/100,000)` =
        ifelse(Population > 0, (`Overall (n)` / Population) * 1e5, NA_real_)
    ) %>%
    left_join(aamr_by_year, by = "Year") %>%
    mutate(Year = as.character(Year)) %>%
    complete(
      Year = as.character(year_min:year_max),
      fill = list(
        `Overall (n)` = NA_real_,
        `Women (n)` = NA_real_,
        `Men (n)` = NA_real_,
        `American Indian or Alaska Native (n)` = NA_real_,
        `Asian (n)` = NA_real_,
        `Native Hawaiian or Other Pacific Islander (n)` = NA_real_,
        `Black or African American (n)` = NA_real_,
        `White (n)` = NA_real_,
        `More than one race (n)` = NA_real_,
        Population = NA_real_,
        `Crude rate (/100,000)` = NA_real_,
        `AAMR (/100,000), age-adjusted (all ages)` = NA_real_
      )
    ) %>%
    arrange(as.integer(Year))

  if (!add_total_row) {
    return(out %>%
      mutate(
        `Crude rate (/100,000)` = round(`Crude rate (/100,000)`, 2),
        `AAMR (/100,000), age-adjusted (all ages)` =
          round(`AAMR (/100,000), age-adjusted (all ages)`, 2)
      ))
  }

  # ============================================================
  # Total row (pooled across all years; correct AAMR)
  # ============================================================
  total_pop <- ya %>%
    summarise(Population = sum(pop, na.rm = TRUE), .groups = "drop") %>%
    pull(Population)

  total_counts <- base %>%
    summarise(
      Year = "Total",
      `Overall (n)` = sum(Deaths, na.rm = TRUE),
      `Women (n)`   = sum(Deaths[Sex == "Female"], na.rm = TRUE),
      `Men (n)`     = sum(Deaths[Sex == "Male"],   na.rm = TRUE),

      `American Indian or Alaska Native (n)` =
        sum(Deaths[race_unified == "American Indian or Alaska Native"], na.rm = TRUE),
      `Asian (n)` =
        sum(Deaths[race_unified == "Asian"], na.rm = TRUE),
      `Native Hawaiian or Other Pacific Islander (n)` =
        sum(Deaths[race_unified == "Native Hawaiian or Other Pacific Islander"], na.rm = TRUE),
      `Black or African American (n)` =
        sum(Deaths[race_unified == "Black or African American"], na.rm = TRUE),
      `White (n)` =
        sum(Deaths[race_unified == "White"], na.rm = TRUE),
      `More than one race (n)` =
        sum(Deaths[race_unified == "More than one race"], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      Population = total_pop,
      `Crude rate (/100,000)` =
        ifelse(Population > 0, (`Overall (n)` / Population) * 1e5, NA_real_)
    )

  total_aamr <- ya %>%
    group_by(std_age) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(std2000_allages, by = "std_age") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    summarise(
      `AAMR (/100,000), age-adjusted (all ages)` = sum(age_rate * weight, na.rm = TRUE)
    )

  total_row <- total_counts %>% bind_cols(total_aamr)

  # Bind + round + order
  bind_rows(out, total_row) %>%
    mutate(
      `Crude rate (/100,000)` = round(`Crude rate (/100,000)`, 2),
      `AAMR (/100,000), age-adjusted (all ages)` =
        round(`AAMR (/100,000), age-adjusted (all ages)`, 2),
      Year = factor(Year, levels = c(as.character(year_min:year_max), "Total"))
    ) %>%
    arrange(Year)
}

# ============================================================
# Optional: quick diagnostic helper for 2018+ undercount (suppression)
# ============================================================
diagnose_suppression <- function(df, year_min = 2018, year_max = 2025) {
  suppressPackageStartupMessages({
    library(dplyr); library(readr)
  })
  parse_pop_safe <- function(x) {
    x <- na_if(as.character(x), "Not Applicable")
    readr::parse_number(x, na = c("", "NA", "Not Applicable"))
  }
  df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      Year       = suppressWarnings(as.integer(`Year Code`)),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population)
    ) %>%
    filter(Year >= year_min, Year <= year_max) %>%
    summarise(
      rows = n(),
      deaths_sum = sum(Deaths, na.rm = TRUE),
      pop_max_sum = sum(tapply(Population, Year, max, na.rm = TRUE), na.rm = TRUE)
    )
}

```

```{r}
table_N17_aamr <-  make_year_table_with_aamr(
  df = N17,
  subchapter_filter = "Renal failure",
  add_total_row = TRUE
)
table_N17_aamr
```

```{r}
table_A40_aamr <- make_year_table_with_aamr(
  df = A40,
  subchapter_filter = "Other bacterial diseases",
  add_total_row = TRUE
)
table_A40_aamr 
```

```{r}
table_N17A40_renal_aamr <- make_year_table_with_aamr(
  df = N17A40_RF,
  add_total_row = TRUE
)
table_N17A40_renal_aamr
```
```{r}

```

## Period Map

```{r}
pod <- read_csv("Data/pod.csv", show_col_types = FALSE)

pod_wide <- pod %>%
  select(period, dataset, place_group, total_deaths) %>%
  mutate(
    place_group = ifelse(place_group == "Medical Facility",
                         "Avg annual deaths (Medical Facility)",
                         "Avg annual deaths (Non-medical facility)")
  ) %>%
  pivot_wider(
    names_from = place_group,
    values_from = total_deaths
  )
```

```{r}
make_period_table_aamr <- function(df,
                                   outcome_name = "ARF-related deaths",
                                   year_min = 2010,
                                   year_max = 2025,
                                   subchapter_filter = NULL) {

  suppressPackageStartupMessages({
    library(dplyr)
    library(tidyr)
    library(tibble)
    library(readr)
  })

  # ============================================================
  # Period definitions
  # ============================================================
  period_map <- tibble(
    period_id = 1:4,
    period = c("Pre-pandemic (2010–2019)",
               "Pandemic (2020–2022)",
               "Transition (2023)",
               "Post-pandemic (2024–2025)"),
    year_start = c(2010, 2020, 2023, 2024),
    year_end   = c(2019, 2022, 2023, 2025)
  )

  add_period <- function(x) {
    x %>%
      left_join(period_map, join_by(Year >= year_start, Year <= year_end)) %>%
      filter(!is.na(period))
  }

  crude_rate <- function(deaths, pop) {
    ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)
  }

  # ============================================================
  # 2000 US standard population (AgeAdj Table II)
  # Expanded to Five-Year Age Groups (ALL ages)
  # ============================================================
  std2000_allages <- tibble(
    age_code = c(
      "1","1-4",
      "5-9","10-14",
      "15-19","20-24",
      "25-29","30-34",
      "35-39","40-44",
      "45-49","50-54",
      "55-59","60-64",
      "65-69","70-74",
      "75-79","80-84",
      "85+"
    ),
    std_million = c(
      13818, 55317,
      145565/2, 145565/2,
      138646/2, 138646/2,
      135573/2, 135573/2,
      162613/2, 162613/2,
      134834/2, 134834/2,
      87247/2,  87247/2,
      66037/2,  66037/2,
      44842/2,  44842/2,
      15508
    )
  ) %>%
    mutate(
      std_million = round(std_million),
      weight_all = std_million / sum(std_million)
    )

  # ============================================================
  # Broad age bins (your exact spec)
  # ============================================================
  recode_to_std_age <- function(age_code) {
    dplyr::case_when(
      age_code %in% c("85-89", "90-94", "95-99", "100+") ~ "85+",
      TRUE ~ age_code
    )
  }

  # ============================================================
  # Broad age bins (your exact spec)
  # NOTE: keep the detailed 85+ bins here; they will be recoded to 85+
  # for standardization, but still belong to 65+.
  # ============================================================
  age_group_map <- list(
    "<20"   = c("1","1-4","5-9","10-14","15-19"),
    "20–44" = c("20-24","25-29","30-34","35-39","40-44"),
    "45–64" = c("45-49","50-54","55-59","60-64"),
    "65+"   = c("65-69","70-74","75-79","80-84","85-89","90-94","95-99","100+","85+")
  )


  # ============================================================
  # Prep data (CDC WONDER robust parsing)
  # ============================================================
  x <- df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      Year = suppressWarnings(as.integer(`Year Code`)),
      Deaths = suppressWarnings(as.numeric(Deaths)),
      Population = readr::parse_number(
        na_if(as.character(Population), "Not Applicable")
      )
    ) %>%
    filter(Year >= year_min, Year <= year_max)

  if (!is.null(subchapter_filter)) {
    x <- x %>% filter(`MCD - ICD Sub-Chapter` %in% subchapter_filter)
  }

  x <- x %>%
    mutate(
      race_source = coalesce(`Single Race 6`, Race),
      race_unified = case_when(
        race_source == "American Indian or Alaska Native" ~ "American Indian or Alaska Native",
        race_source %in% c("Asian", "Asian or Pacific Islander") ~ "Asian",
        race_source == "Native Hawaiian or Other Pacific Islander" ~ "Native Hawaiian or Other Pacific Islander",
        race_source == "Black or African American" ~ "Black or African American",
        race_source == "White" ~ "White",
        race_source == "More than one race" ~ "More than one race",
        TRUE ~ NA_character_
      ),
      age_code = `Five-Year Age Groups Code`,
      age_cat = case_when(
        age_code %in% age_group_map[["<20"]]   ~ "<20",
        age_code %in% age_group_map[["20–44"]] ~ "20–44",
        age_code %in% age_group_map[["45–64"]] ~ "45–64",
        age_code %in% age_group_map[["65+"]]   ~ "65+",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(race_unified)) %>%   # drop "Not available"
    add_period()

  # ============================================================
  # Helper: overall AAMR across ALL ages (fixed full weights)
  # ============================================================
  calc_aamr_allages <- function(data, group_vars) {
    data %>%
      filter(age_code %in% std2000_allages$age_code) %>%
      group_by(across(all_of(group_vars)), age_code) %>%
      summarise(
        deaths = sum(Deaths, na.rm = TRUE),
        pop    = sum(Population, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      left_join(std2000_allages, by = "age_code") %>%
      mutate(age_rate = ifelse(pop > 0, deaths / pop * 1e5, NA_real_)) %>%
      group_by(across(all_of(group_vars))) %>%
      summarise(
        AAMR_allages = sum(age_rate * weight_all, na.rm = TRUE),
        .groups = "drop"
      )
  }

  # ============================================================
  # Helper: AAMR WITHIN a broad age bin (weights re-normalized)
  # ============================================================
  calc_aamr_within_agebin <- function(data, group_vars, age_codes) {

    std_sub <- std2000_allages %>%
      filter(age_code %in% age_codes) %>%
      mutate(weight = std_million / sum(std_million))

    data %>%
      filter(age_code %in% age_codes) %>%
      group_by(across(all_of(group_vars)), age_code) %>%
      summarise(
        deaths = sum(Deaths, na.rm = TRUE),
        pop    = sum(Population, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      left_join(std_sub, by = "age_code") %>%
      mutate(age_rate = ifelse(pop > 0, deaths / pop * 1e5, NA_real_)) %>%
      group_by(across(all_of(group_vars))) %>%
      summarise(
        AAMR = sum(age_rate * weight, na.rm = TRUE),
        .groups = "drop"
      )
  }

  # ============================================================
  # Period totals: deaths/pop + crude
  # ============================================================
  period_totals <- x %>%
    group_by(period_id, period) %>%
    summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      population = sum(Population, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(crude = crude_rate(deaths, population))

  period_aamr <- calc_aamr_allages(x, c("period_id", "period"))
  
  sex_totals <- x %>%
    group_by(period_id, period, Sex) %>%
    summarise(
      deaths_sex = sum(Deaths, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(period_totals %>% select(period_id, period, deaths), by = c("period_id", "period")) %>%
    mutate(
      pct_sex = ifelse(deaths > 0, deaths_sex / deaths * 100, NA_real_)
    ) %>%
    select(period_id, period, Sex, deaths_sex, pct_sex) %>%
    pivot_wider(
      names_from = Sex,
      values_from = c(deaths_sex, pct_sex),
      names_sep = "_"
    )

  # ============================================================
  # Sex AAMR (overall all-ages)
  # ============================================================
  sex_aamr <- calc_aamr_allages(x, c("period_id", "period", "Sex")) %>%
    select(period_id, period, Sex, AAMR_allages) %>%
    pivot_wider(names_from = Sex, values_from = AAMR_allages) %>%
    rename(Male = Male, Female = Female)

  # ============================================================
  # ✅ AGE-BIN AAMR (this is what you asked for)
  # One AAMR per broad age bin (<20, 20–44, 45–64, 65+) per period
  # ============================================================
  age_aamr <- bind_rows(
    calc_aamr_within_agebin(x, c("period_id", "period"), age_group_map[["<20"]])   %>% mutate(age_cat = "<20"),
    calc_aamr_within_agebin(x, c("period_id", "period"), age_group_map[["20–44"]]) %>% mutate(age_cat = "20–44"),
    calc_aamr_within_agebin(x, c("period_id", "period"), age_group_map[["45–64"]]) %>% mutate(age_cat = "45–64"),
    calc_aamr_within_agebin(x, c("period_id", "period"), age_group_map[["65+"]])   %>% mutate(age_cat = "65+")
  ) %>%
    select(period_id, period, age_cat, AAMR) %>%
    pivot_wider(names_from = age_cat, values_from = AAMR)

  # ============================================================
  # Race AAMR (overall all-ages)
  # ============================================================
  race_aamr <- calc_aamr_allages(x, c("period_id", "period", "race_unified")) %>%
    select(period_id, period, race_unified, AAMR_allages) %>%
    pivot_wider(names_from = race_unified, values_from = AAMR_allages)

  # ============================================================
  # Assemble numeric table
  # ============================================================
  period_table <- period_totals %>%
    left_join(period_aamr, by = c("period_id", "period")) %>%
    left_join(sex_totals,   by = c("period_id", "period")) %>% 
    left_join(sex_aamr,    by = c("period_id", "period")) %>%
    left_join(age_aamr,    by = c("period_id", "period")) %>%
    left_join(race_aamr,   by = c("period_id", "period")) %>%
    arrange(period_id) %>%
    transmute(
      period_id, period,
      `Total deaths, n` = deaths,
      `Population` = population,
      `Crude mortality rate (/100,000)` = crude,
      `AAMR (/100,000), age-adjusted (all ages)` = AAMR_allages,
      `Male deaths, n`   = deaths_sex_Male,
      `Male deaths, %`   = pct_sex_Male,
      `Female deaths, n` = deaths_sex_Female,
      `Female deaths, %` = pct_sex_Female,
      `Male AAMR (/100,000)` = Male,
      `Female AAMR (/100,000)` = Female,
      `<20 AAMR (/100,000)` = `<20`,
      `20–44 AAMR (/100,000)` = `20–44`,
      `45–64 AAMR (/100,000)` = `45–64`,
      `65+ AAMR (/100,000)` = `65+`,
      `American Indian or Alaska Native AAMR (/100,000)` = `American Indian or Alaska Native`,
      `Asian AAMR (/100,000)` = Asian,
      `Native Hawaiian or Other Pacific Islander AAMR (/100,000)` = `Native Hawaiian or Other Pacific Islander`,
      `Black or African American AAMR (/100,000)` = `Black or African American`,
      `White AAMR (/100,000)` = White,
      `More than one race AAMR (/100,000)` = `More than one race`
    )

  # ============================================================
  # Publication formatting
  # ============================================================
   pub <- period_table %>%
    select(-period_id) %>%
    pivot_longer(cols = -period, names_to = "Metric", values_to = "Value") %>%
    mutate(
      Value = dplyr::case_when(
        Metric %in% c("Total deaths, n", "Population",
                      "Male deaths, n", "Female deaths, n") ~
          formatC(round(Value, 0), format = "f", big.mark = ",", digits = 0),

        Metric %in% c("Male deaths, %", "Female deaths, %") ~
          formatC(round(Value, 1), format = "f", digits = 1),

        TRUE ~
          formatC(round(Value, 4), format = "f", big.mark = ",", digits = 4)
      )
    ) %>%
    pivot_wider(names_from = period, values_from = Value)

  # ---- ✅ Build n (%) rows using the already-formatted n and % rows ----
  male_n_pct <- pub %>%
    filter(Metric %in% c("Male deaths, n", "Male deaths, %")) %>%
    pivot_longer(cols = -Metric, names_to = "period", values_to = "val") %>%
    mutate(kind = ifelse(Metric == "Male deaths, n", "n", "%")) %>%
    select(-Metric) %>%
    pivot_wider(names_from = kind, values_from = val) %>%
    mutate(Metric = "Male deaths, n (%)",
           val = paste0(n, " (", `%`, "%)")) %>%
    select(Metric, period, val) %>%
    pivot_wider(names_from = period, values_from = val)

  female_n_pct <- pub %>%
    filter(Metric %in% c("Female deaths, n", "Female deaths, %")) %>%
    pivot_longer(cols = -Metric, names_to = "period", values_to = "val") %>%
    mutate(kind = ifelse(Metric == "Female deaths, n", "n", "%")) %>%
    select(-Metric) %>%
    pivot_wider(names_from = kind, values_from = val) %>%
    mutate(Metric = "Female deaths, n (%)",
           val = paste0(n, " (", `%`, "%)")) %>%
    select(Metric, period, val) %>%
    pivot_wider(names_from = period, values_from = val)

  # ✅ Insert the n(%) rows right after Total deaths row (nice for Table 1 style)
  pub <- bind_rows(
    pub %>% filter(Metric == "Total deaths, n"),
    male_n_pct,
    female_n_pct,
    pub %>% filter(!Metric %in% c("Total deaths, n", "Male deaths, n", "Male deaths, %", "Female deaths, n", "Female deaths, %"))
  )
    list(
    outcome = outcome_name,
    numeric_period_table = period_table,
    publication_table = pub
  )
}


```

```{r}
# Renal failure (ARF) period table with AAMR
arf_period <- make_period_table_aamr(N17, subchapter_filter = "Renal failure")
arf_period$publication_table

# Sepsis period table with AAMR (adjust subchapter_filter to your definition)
sepsis_period <- make_period_table_aamr(A40, subchapter_filter = "Other bacterial diseases")
sepsis_period$publication_table

# Combined dataset but renal-failure-only
renal_only_from_combined <- make_period_table_aamr(N17A40, subchapter_filter = "Renal failure")
renal_only_from_combined$publication_table

```

```{r}
add_pod_using_period_totals <- function(period_list_obj, pod_total_long, pod_dataset_name = NULL) {

  suppressPackageStartupMessages({
    library(dplyr)
    library(tidyr)
    library(stringr)
    library(readr)
  })

  out_name <- if (!is.null(pod_dataset_name)) pod_dataset_name else period_list_obj$outcome

  # ---- 1) Get TOTAL deaths per period from make_period_table_aamr output ----
  # Prefer numeric_period_table if present (numeric), else parse from publication_table.
  if (!is.null(period_list_obj$numeric_period_table)) {

    totals_by_period <- period_list_obj$numeric_period_table %>%
      transmute(
        period,
        total_deaths = `Total deaths, n`
      )

  } else {
    # publication_table shape: Metric row + period columns (strings with commas)
    totals_by_period <- period_list_obj$publication_table %>%
      filter(Metric == "Total deaths, n") %>%
      pivot_longer(
        cols = -Metric,
        names_to = "period",
        values_to = "total_deaths_chr"
      ) %>%
      transmute(
        period,
        total_deaths = readr::parse_number(total_deaths_chr)
      )
  }

  # ---- 2) Get MEDICAL deaths per period from pod_total_long ----
  med_by_period <- pod_total_long %>%
    filter(dataset == out_name) %>%
    mutate(
      place_group = case_when(
        str_detect(place_group, "^Medical Facility") ~ "Medical Facility",
        str_detect(place_group, "^Non-medical")      ~ "Non-medical facility",
        TRUE ~ as.character(place_group)
      )
    ) %>%
    filter(place_group == "Medical Facility") %>%
    group_by(period) %>%
    summarise(medical_deaths = sum(total_deaths, na.rm = TRUE), .groups = "drop")

  if (nrow(med_by_period) == 0) {
    warning("No Medical Facility POD rows found for dataset == '", out_name, "'.")
    return(period_list_obj)
  }

  # ---- 3) Combine + compute Non-medical = Total - Medical + n(%) ----
  pod_calc <- totals_by_period %>%
    left_join(med_by_period, by = "period") %>%
    mutate(
      medical_deaths = coalesce(medical_deaths, 0),
      nonmed_deaths  = pmax(total_deaths - medical_deaths, 0),
      med_pct   = ifelse(total_deaths > 0, 100 * medical_deaths / total_deaths, NA_real_),
      nonmed_pct= ifelse(total_deaths > 0, 100 * nonmed_deaths  / total_deaths, NA_real_)
    )

  pod_pub <- bind_rows(
    pod_calc %>%
      transmute(
        period,
        Metric = "Place of death: Medical Facility",
        Value = paste0(
          formatC(round(medical_deaths, 0), format = "f", digits = 0, big.mark = ","),
          " (",
          formatC(med_pct, format = "f", digits = 1),
          "%)"
        )
      ),
    pod_calc %>%
      transmute(
        period,
        Metric = "Place of death: Non-medical facility",
        Value = paste0(
          formatC(round(nonmed_deaths, 0), format = "f", digits = 0, big.mark = ","),
          " (",
          formatC(nonmed_pct, format = "f", digits = 1),
          "%)"
        )
      )
  ) %>%
    distinct(period, Metric, .keep_all = TRUE) %>%
    pivot_wider(names_from = period, values_from = Value)

  # ---- 4) Replace any existing POD rows, then append ----
  period_list_obj$publication_table <- period_list_obj$publication_table %>%
    filter(!str_detect(Metric, "^Place of death:")) %>%
    bind_rows(pod_pub)

  period_list_obj
}

# ============================================================
# USE (match pod dataset names)
# pod_total_long is your tot_all/pod table with: period, place_group, total_deaths, dataset
# ============================================================

arf_period <- add_pod_using_period_totals(arf_period, pod, pod_dataset_name = "AKI")
sepsis_period <- add_pod_using_period_totals(sepsis_period, pod, pod_dataset_name = "Sepsis")
renal_only_from_combined <- add_pod_using_period_totals(renal_only_from_combined, pod, pod_dataset_name = "AKI & Sepsis")

arf_period$publication_table
sepsis_period$publication_table
renal_only_from_combined$publication_table
```

## Age

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(readr)
})

# ============================================================
# 2000 U.S. standard population weights (ALL ages; 5-year bins)
# ============================================================
std2000_allages <- tibble(
  age_code = c(
    "1","1-4",
    "5-9","10-14",
    "15-19","20-24",
    "25-29","30-34",
    "35-39","40-44",
    "45-49","50-54",
    "55-59","60-64",
    "65-69","70-74",
    "75-79","80-84",
    "85+"
  ),
  std_million = c(
    13818, 55317,
    145565/2, 145565/2,
    138646/2, 138646/2,
    135573/2, 135573/2,
    162613/2, 162613/2,
    134834/2, 134834/2,
    87247/2,  87247/2,
    66037/2,  66037/2,
    44842/2,  44842/2,
    15508
  )
) %>%
  mutate(
    std_million = round(std_million),
    weight = std_million / sum(std_million)
  )

# ============================================================
# Recode old-age detail bins into 85+ (so they are kept)
# ============================================================
recode_to_std_age <- function(age_code) {
  age_code <- as.character(age_code)
  dplyr::case_when(
    age_code %in% c("85-89", "90-94", "95-99", "100+") ~ "85+",
    TRUE ~ age_code
  )
}

# ============================================================
# Age-band map
# ============================================================
age_band_map <- tibble(
  age_code = std2000_allages$age_code,
  age4 = c(
    rep("<20",    5),
    rep("20_44",  5),
    rep("45_64",  4),
    rep("65_plus",5)
  )
)

# ============================================================
# Safe population parsing
# ============================================================
parse_pop_safe <- function(x) {
  x <- na_if(as.character(x), "Not Applicable")
  readr::parse_number(x, na = c("", "NA", "Not Applicable"))
}

# ============================================================
# Partition var logic aligned with your collapse_for_rates():
# - prefer Single Race 6 over Race
# - partition by Sex + (Single Race 6 or Race), when present
# ============================================================
race_part_var <- function(dat) {
  if ("Single Race 6" %in% names(dat)) return("Single Race 6")
  if ("Race" %in% names(dat))          return("Race")
  return(NULL)
}

# ============================================================
# Collapse YearKey × age_code with correct pop handling
# (ALIGNED to collapse_for_rates for overall/sex pipelines)
# 1) group by YearKey × age_code × partition vars
# 2) pop = max(Population) within stratum
# 3) sum across strata to YearKey × age_code
# ============================================================
collapse_year_age <- function(dat) {
  rp <- race_part_var(dat)

  pv <- c("Sex", rp)
  pv <- pv[!is.null(pv)]
  pv <- pv[pv %in% names(dat)]

  dat %>%
    group_by(across(all_of(c("YearKey", "age_code", pv)))) %>%
    summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(pop = ifelse(is.infinite(pop), NA_real_, pop)) %>%
    group_by(YearKey, age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )
}

# ============================================================
# Band-specific AAMR by YearKey (2010–2025)
# - deduped pop via collapse_year_age()
# - renormalize weights within each band
# ============================================================
calc_band_aamr_by_yearcode <- function(df, year_min = 2010, year_max = 2025) {

  base <- df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      YearCode   = suppressWarnings(as.integer(`Year Code`)),
      YearKey    = as.character(YearCode),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code   = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    filter(!is.na(YearCode), YearCode >= year_min, YearCode <= year_max) %>%
    filter(age_code %in% std2000_allages$age_code)

  ya <- collapse_year_age(base)

  ya %>%
    left_join(age_band_map, by = "age_code") %>%
    left_join(std2000_allages %>% select(age_code, weight), by = "age_code") %>%
    group_by(YearKey, age4) %>%
    mutate(
      w_band   = weight / sum(weight, na.rm = TRUE),
      age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)
    ) %>%
    summarise(
      AAMR_band = sum(age_rate * w_band, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    complete(
      YearKey = as.character(year_min:year_max),
      age4    = c("<20", "20_44", "45_64", "65_plus"),
      fill    = list(AAMR_band = NA_real_)
    ) %>%
    arrange(as.integer(YearKey), factor(age4, levels = c("<20","20_44","45_64","65_plus")))
}

# ============================================================
# Total row for band-AAMR (YearKey = "Total")  -- FIXED
# IMPORTANT: pool across years to age_code before weighting
# ============================================================
calc_band_aamr_total_yearcode <- function(df, year_min = 2010, year_max = 2025) {

  base <- df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      YearCode   = suppressWarnings(as.integer(`Year Code`)),
      YearKey    = as.character(YearCode),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code   = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    filter(!is.na(YearCode), YearCode >= year_min, YearCode <= year_max) %>%
    filter(age_code %in% std2000_allages$age_code)

  ya_year <- collapse_year_age(base)

  # pool across years -> age_code totals
  ya_total <- ya_year %>%
    group_by(age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  ya_total %>%
    left_join(age_band_map, by = "age_code") %>%
    left_join(std2000_allages %>% select(age_code, weight), by = "age_code") %>%
    group_by(age4) %>%
    mutate(
      w_band   = weight / sum(weight, na.rm = TRUE),
      age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)
    ) %>%
    summarise(
      YearKey   = "Total",
      AAMR_band = sum(age_rate * w_band, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(age4 = factor(age4, levels = c("<20","20_44","45_64","65_plus"))) %>%
    arrange(age4)
}

# ============================================================
# Final wrapper: deaths by band + band-AAMR by YearKey + Total
# ============================================================
year_age_deaths_band_aamr <- function(df, prefix, year_min = 2010, year_max = 2025) {

  # ---- deaths by band (counts don't need population dedupe) ----
  by_year_age <- df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      YearCode = suppressWarnings(as.integer(`Year Code`)),
      YearKey  = as.character(YearCode),
      Deaths   = suppressWarnings(as.numeric(Deaths)),
      age_code = recode_to_std_age(`Five-Year Age Groups Code`),
      age4 = case_when(
        age_code %in% c("1","1-4","5-9","10-14","15-19") ~ "<20",
        age_code %in% c("20-24","25-29","30-34","35-39","40-44") ~ "20_44",
        age_code %in% c("45-49","50-54","55-59","60-64") ~ "45_64",
        age_code %in% c("65-69","70-74","75-79","80-84","85+") ~ "65_plus",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(YearCode), YearCode >= year_min, YearCode <= year_max, !is.na(age4)) %>%
    group_by(YearKey, age4) %>%
    summarise(deaths = sum(Deaths, na.rm = TRUE), .groups = "drop") %>%
    complete(
      YearKey = as.character(year_min:year_max),
      age4    = c("<20", "20_44", "45_64", "65_plus"),
      fill    = list(deaths = 0)
    )

  total_deaths_rows <- by_year_age %>%
    group_by(age4) %>%
    summarise(YearKey = "Total", deaths = sum(deaths, na.rm = TRUE), .groups = "drop")

  deaths_wide <- bind_rows(by_year_age, total_deaths_rows) %>%
    pivot_wider(names_from = age4, values_from = deaths, values_fill = 0) %>%
    rename(
      !!paste0(prefix, "_deaths_lt20")    := `<20`,
      !!paste0(prefix, "_deaths_20_44")   := `20_44`,
      !!paste0(prefix, "_deaths_45_64")   := `45_64`,
      !!paste0(prefix, "_deaths_65_plus") := `65_plus`
    )

  # ---- band-AAMR by YearKey + Total ----
  aamr_year  <- calc_band_aamr_by_yearcode(df, year_min, year_max)
  aamr_total <- calc_band_aamr_total_yearcode(df, year_min, year_max)

  aamr_all <- bind_rows(aamr_year, aamr_total)

  rates_wide <- aamr_all %>%
    mutate(age4 = as.character(age4)) %>%
    pivot_wider(names_from = age4, values_from = AAMR_band) %>%
    rename(
      !!paste0(prefix, "_<20")    := `<20`,
      !!paste0(prefix, "_20_44")  := `20_44`,
      !!paste0(prefix, "_45_64")  := `45_64`,
      !!paste0(prefix, "_65+")    := `65_plus`
    )

  list(deaths = deaths_wide, rates = rates_wide)
}

```

```{r}

# ---- Build for the 3 datasets ----
arf   <- year_age_deaths_band_aamr(N17, "AKI")
sepsis  <- year_age_deaths_band_aamr(A40, "Sepsis")
both <- year_age_deaths_band_aamr(N17A40_RF, "AKI_Sepsis")

# ---- Combine into one deaths table ----
deaths_table_age <- arf$deaths %>%
  full_join(sepsis$deaths, by = "YearKey") %>%
  full_join(both$deaths,   by = "YearKey") %>%
  mutate(YearKey = factor(YearKey, levels = c(as.character(2010:2025), "Total"))) %>%
  arrange(YearKey)

# ---- Combine into one crude-rate table ----
rate_table_age <- arf$rates %>%
  full_join(sepsis$rates, by = "YearKey") %>%
  full_join(both$rates,   by = "YearKey") %>%
  mutate(YearKey = factor(YearKey, levels = c(as.character(2010:2025), "Total"))) %>%
  arrange(YearKey) %>%
  # Optional: round rates nicely
  mutate(across(where(is.numeric), ~ round(.x, 2)))

deaths_table_age
rate_table_age

```

## Sex

```{r}
# ============================================================
# YEAR x SEX tables ONLY (delete age bands)
# - Deaths by Year x Sex (+ Overall)
# - AAMR by Year x Sex (+ Overall), using ALL-AGES standardization
# - Also adds a "Total" row (2010–2025 combined)
# Uses Year Code for time.
# ============================================================

# ------------------------------------------------------------
# 0) Population parser (safer)
# ------------------------------------------------------------
parse_pop_safe <- function(x) {
  x <- na_if(as.character(x), "Not Applicable")
  readr::parse_number(x, na = c("", "NA", "Not Applicable"))
}

# ------------------------------------------------------------
# helper: choose race partition column like your other pipeline
# ------------------------------------------------------------
race_part_var <- function(dat) {
  if ("Single Race 6" %in% names(dat)) return("Single Race 6")
  if ("Race" %in% names(dat))          return("Race")
  return(NULL)
}

# ============================================================
# AAMR by Year x Sex (ALL ages; standard weights)
# + Overall per year (recomputed)
# + Total (pooled 2010–2025; recomputed)
# ============================================================
calc_aamr_year_sex <- function(df, year_min = 2010, year_max = 2025) {

  base <- df %>%
    dplyr::filter(is.na(Notes) | Notes != "Total") %>%
    dplyr::mutate(
      YearCode    = suppressWarnings(as.integer(`Year Code`)),
      YearKey     = as.character(YearCode),
      Deaths      = suppressWarnings(as.numeric(Deaths)),
      Population  = parse_pop_safe(Population),
      age_code    = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    dplyr::filter(!is.na(YearCode), YearCode >= year_min, YearCode <= year_max) %>%
    dplyr::filter(age_code %in% std2000_allages$age_code) %>%
    dplyr::filter(!is.na(Sex))

  # ---- partition vars to avoid pop double-counting ----
  rp <- race_part_var(base)
  pv <- c("Sex", rp)
  pv <- pv[!is.null(pv)]
  pv <- pv[pv %in% names(base)]

  # ------------------------------------------------------------
  # 1) Collapse away duplicated population correctly:
  # 1a) Year×Sex×Age×(race partition) : pop = max(pop) (dedupe repeats)
  # 1b) then SUM across race partition to Year×Sex×Age
  # ------------------------------------------------------------
  sex_age_year <- base %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c("YearKey", "YearCode", "age_code", pv)))) %>%
    dplyr::summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::mutate(pop = dplyr::if_else(is.infinite(pop), NA_real_, pop)) %>%
    dplyr::group_by(YearKey, YearCode, Sex, age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  # ------------------------------------------------------------
  # 2) Sex-specific AAMR by year
  # ------------------------------------------------------------
  sex_year <- sex_age_year %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::mutate(age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    dplyr::group_by(YearKey, YearCode, Sex) %>%
    dplyr::summarise(
      AAMR = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    )

  # ------------------------------------------------------------
  # 3) Overall AAMR by year (both sexes pooled at age level)
  # ------------------------------------------------------------
  overall_year <- sex_age_year %>%
    dplyr::group_by(YearKey, YearCode, age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::mutate(age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    dplyr::group_by(YearKey, YearCode) %>%
    dplyr::summarise(
      AAMR = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(Sex = "Overall")

  out_year <- dplyr::bind_rows(sex_year, overall_year) %>%
    dplyr::mutate(
      Sex = trimws(as.character(Sex)),
      Sex = factor(Sex, levels = c("Male", "Female", "Overall"))
    ) %>%
    tidyr::complete(
      YearKey = as.character(year_min:year_max),
      Sex = factor(c("Male", "Female", "Overall"), levels = c("Male","Female","Overall")),
      fill = list(AAMR = NA_real_, YearCode = NA_integer_)
    ) %>%
    dplyr::mutate(YearCode = suppressWarnings(as.integer(YearKey))) %>%
    dplyr::select(YearKey, Sex, AAMR)

  # ------------------------------------------------------------
  # 4) Total row (2010–2025): Sex-specific
  # Pool across years at age level first
  # ------------------------------------------------------------
  total_sex <- sex_age_year %>%
    dplyr::group_by(Sex, age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::mutate(age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    dplyr::group_by(Sex) %>%
    dplyr::summarise(
      YearKey = "Total",
      AAMR = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    )

  # ------------------------------------------------------------
  # 5) Total row: Overall (pool sexes at age level)
  # ------------------------------------------------------------
  total_overall <- sex_age_year %>%
    dplyr::group_by(age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::mutate(age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    dplyr::summarise(
      YearKey = "Total",
      Sex = "Overall",
      AAMR = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    )

  dplyr::bind_rows(out_year, total_sex, total_overall) %>%
    dplyr::mutate(Sex = factor(as.character(Sex), levels = c("Male","Female","Overall"))) %>%
    dplyr::arrange(dplyr::if_else(YearKey == "Total", 9999L, suppressWarnings(as.integer(YearKey))),
                  Sex)
}

# ============================================================
# Main: deaths by (YearKey x Sex) + AAMR by same
# Output is WIDE with columns like:
# AKI_deaths_Male, AKI_AAMR_Overall, etc.
# ============================================================
year_sex_deaths_aamr <- function(df, prefix, year_min = 2010, year_max = 2025) {

  deaths_long <- df %>%
    dplyr::filter(is.na(Notes) | Notes != "Total") %>%
    dplyr::mutate(
      YearCode = suppressWarnings(as.integer(`Year Code`)),
      YearKey  = as.character(YearCode),
      Deaths   = suppressWarnings(as.numeric(Deaths))
    ) %>%
    dplyr::filter(!is.na(YearCode), YearCode >= year_min, YearCode <= year_max) %>%
    dplyr::filter(!is.na(Sex)) %>%
    dplyr::group_by(YearKey, Sex) %>%
    dplyr::summarise(deaths = sum(Deaths, na.rm = TRUE), .groups = "drop") %>%
    dplyr::mutate(Sex = trimws(as.character(Sex))) %>%
    dplyr::filter(Sex %in% c("Male","Female")) %>%
    tidyr::complete(
      YearKey = as.character(year_min:year_max),
      Sex = c("Male", "Female"),
      fill = list(deaths = 0)
    )

  deaths_overall_year <- deaths_long %>%
    dplyr::group_by(YearKey) %>%
    dplyr::summarise(Sex = "Overall", deaths = sum(deaths, na.rm = TRUE), .groups = "drop")

  deaths_long2 <- dplyr::bind_rows(deaths_long, deaths_overall_year)

  deaths_total <- deaths_long2 %>%
    dplyr::group_by(Sex) %>%
    dplyr::summarise(YearKey = "Total", deaths = sum(deaths, na.rm = TRUE), .groups = "drop")

  deaths_long2 <- dplyr::bind_rows(deaths_long2, deaths_total)

  deaths_wide <- deaths_long2 %>%
    tidyr::pivot_wider(names_from = Sex, values_from = deaths, values_fill = 0) %>%
    dplyr::rename_with(~ paste0(prefix, "_deaths_", .x), -YearKey)

  # ---- AAMR long + wide ----
  aamr_long <- calc_aamr_year_sex(df, year_min, year_max)

  rates_wide <- aamr_long %>%
    tidyr::pivot_wider(names_from = Sex, values_from = AAMR) %>%
    dplyr::rename_with(~ paste0(prefix,"_AAMR_", .x), -YearKey)

  list(deaths = deaths_wide, rates = rates_wide)
}
# ============================================================
# Build for three datasets
# ============================================================
arf    <- year_sex_deaths_aamr(N17,    "AKI")
sepsis <- year_sex_deaths_aamr(A40,    "Sepsis")
both   <- year_sex_deaths_aamr(N17A40_RF, "AKI_Sepsis")

# ============================================================
# Join tables (use YearKey, not Year)
# ============================================================
deaths_table_sex <- arf$deaths %>%
  full_join(sepsis$deaths, by = "YearKey") %>%
  full_join(both$deaths,   by = "YearKey") %>%
  mutate(YearKey = factor(YearKey, levels = c(as.character(2010:2025), "Total"))) %>%
  arrange(YearKey)

aamr_table_sex <- arf$rates %>%
  full_join(sepsis$rates, by = "YearKey") %>%
  full_join(both$rates,   by = "YearKey") %>%
  mutate(YearKey = factor(YearKey, levels = c(as.character(2010:2025), "Total"))) %>%
  arrange(YearKey) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

deaths_table_sex
aamr_table_sex

```

## Portional

```{r}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(readr)
})

# ----------------------------
# Standard weights (your same table)
# ----------------------------
std2000_allages <- tibble(
  age_code = c(
    "1","1-4","5-9","10-14","15-19","20-24","25-29","30-34","35-39","40-44",
    "45-49","50-54","55-59","60-64","65-69","70-74","75-79","80-84","85+"
  ),
  std_million = c(
    13818, 55317,
    145565/2, 145565/2,
    138646/2, 138646/2,
    135573/2, 135573/2,
    162613/2, 162613/2,
    134834/2, 134834/2,
    87247/2,  87247/2,
    66037/2,  66037/2,
    44842/2,  44842/2,
    15508
  )
) %>%
  mutate(std_million = round(std_million),
         weight = std_million / sum(std_million))

# ----------------------------
# Helpers (shared)
# ----------------------------
parse_pop_safe <- function(x) {
  x <- na_if(as.character(x), "Not Applicable")
  readr::parse_number(x, na = c("", "NA", "Not Applicable"))
}

recode_to_std_age <- function(age_code) {
  dplyr::case_when(
    age_code %in% c("85-89", "90-94", "95-99", "100+") ~ "85+",
    TRUE ~ age_code
  )
}

get_part_vars <- function(dat) {
  pv <- intersect(c("Sex", "Single Race 6", "Race"), names(dat))
  if (all(c("Race", "Single Race 6") %in% pv)) pv <- setdiff(pv, "Race")
  pv
}

# ------------------------------------------------------------
# Core: build Year×Age totals with corrected pop
# (dedupe pop with max() within strata; then sum strata)
# ------------------------------------------------------------
year_age_totals <- function(dat, year_var = "YearKey") {
  pv <- get_part_vars(dat)

  dat %>%
    group_by(across(all_of(c(year_var, "age_code", pv)))) %>%
    summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(pop = ifelse(is.infinite(pop), NA_real_, pop)) %>%
    group_by(across(all_of(c(year_var, "age_code")))) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )
}

# ------------------------------------------------------------
# ✅ POOLED Total AAMR (aligned)
# Uses the same “ya” object used for yearly AAMR:
#   - sum deaths/pop across ALL YEARS within age_code
#   - apply weights
# ------------------------------------------------------------
pooled_total_aamr_from_ya <- function(ya) {
  ya %>%
    group_by(age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(std2000_allages, by = "age_code") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    summarise(AAMR = sum(age_rate * weight, na.rm = TRUE)) %>%
    pull(AAMR)
}

# ============================================================
# 1) Yearly AAMR table (with aligned Total row)
# ============================================================
yearly_aamr_allages <- function(df, label,
                                year_min = 2010, year_max = 2025,
                                subchapter_filter = NULL,
                                add_total_row = TRUE) {

  base <- df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      YearCode   = suppressWarnings(as.integer(`Year Code`)),
      YearKey    = as.character(YearCode),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code   = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    filter(YearCode >= year_min, YearCode <= year_max) %>%
    filter(age_code %in% std2000_allages$age_code)

  if (!is.null(subchapter_filter)) {
    base <- base %>% filter(`MCD - ICD Sub-Chapter` %in% subchapter_filter)
  }

  ya <- year_age_totals(base, year_var = "YearKey")

  out_year <- ya %>%
    left_join(std2000_allages, by = "age_code") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    group_by(YearKey) %>%
    summarise(AAMR = sum(age_rate * weight, na.rm = TRUE), .groups = "drop") %>%
    complete(YearKey = as.character(year_min:year_max), fill = list(AAMR = NA_real_)) %>%
    arrange(as.integer(YearKey)) %>%
    rename(!!label := AAMR)

  if (!add_total_row) return(out_year)

  total_aamr <- pooled_total_aamr_from_ya(ya)

  bind_rows(
    out_year,
    tibble(YearKey = "Total", !!label := total_aamr)
  ) %>%
    mutate(YearKey = factor(YearKey, levels = c(as.character(year_min:year_max), "Total")))
}

# ============================================================
# 2) Ratio table (with aligned Total row + aligned Total ratios)
# ============================================================
make_year_ratio_table <- function(aki_df, sepsis_df, both_df,
                                  year_min = 2010, year_max = 2025,
                                  aki_label = "AKI (AAMR all ages)",
                                  sepsis_label = "Sepsis (AAMR all ages)",
                                  both_label = "AKI & Sepsis (AAMR all ages)",
                                  aki_subchapter = NULL,
                                  sepsis_subchapter = NULL,
                                  both_subchapter = NULL) {

  aki_y  <- yearly_aamr_allages(aki_df,  aki_label,  year_min, year_max, aki_subchapter,  add_total_row = TRUE)
  sep_y  <- yearly_aamr_allages(sepsis_df, sepsis_label, year_min, year_max, sepsis_subchapter, add_total_row = TRUE)
  both_y <- yearly_aamr_allages(both_df, both_label, year_min, year_max, both_subchapter, add_total_row = TRUE)

  tab <- aki_y %>%
    left_join(sep_y,  by = "YearKey") %>%
    left_join(both_y, by = "YearKey") %>%
    mutate(
      `Sepsis among AKI ((3)/(1))` =
        if_else(.data[[aki_label]] > 0, .data[[both_label]] / .data[[aki_label]], NA_real_),
      `AKI among Sepsis ((3)/(2))` =
        if_else(.data[[sepsis_label]] > 0, .data[[both_label]] / .data[[sepsis_label]], NA_real_)
    )

  tab
}
# ---- 1) Get yearly values for each dataset ----
arf_tbl  <- year_overall_aamr_allages(N17,    "AKI")
sep_tbl  <- year_overall_aamr_allages(A40,    "Sepsis")
both_tbl <- year_overall_aamr_allages(N17A40_RF, "AKI_Sepsis")

# ---- 2) Combine + compute overlap metrics using ALL-AGES AAMR ----
overlap_table_aamr_allages <- arf_tbl %>%
  full_join(sep_tbl,  by = "YearKey") %>%
  full_join(both_tbl, by = "YearKey") %>%
  mutate(
    `Sepsis among AKI (AAMR)` =
      ifelse(AKI_AAMR_allages > 0, AKI_Sepsis_AAMR_allages / AKI_AAMR_allages, NA_real_),
    `AKI among Sepsis (AAMR)` =
      ifelse(Sepsis_AAMR_allages > 0, AKI_Sepsis_AAMR_allages / Sepsis_AAMR_allages, NA_real_)
  ) %>%
  select(
    YearKey,
    `AKI (AAMR all ages)` = AKI_AAMR_allages,
    `Sepsis (AAMR all ages)` = Sepsis_AAMR_allages,
    `AKI & Sepsis (AAMR all ages)` = AKI_Sepsis_AAMR_allages,
    `Sepsis among AKI ((3)/(1))` = `Sepsis among AKI (AAMR)`,
    `AKI among Sepsis ((3)/(2))` = `AKI among Sepsis (AAMR)`
  ) %>%
  mutate(
    YearKey = factor(YearKey, levels = c(as.character(2010:2025), "Total")),
    across(where(is.numeric), ~ round(.x, 4))
  ) %>%
  arrange(YearKey)

overlap_table_aamr_allages
```

## Overall AAMR

```{r}
pod <- read_csv("Data/pod.csv")
glimpse(pod)
```

```{r}
# ============================================================
# FULL SCRIPT (copy/paste)
# Goal: Make summary_tbl "AAMR (all ages)" match make_year_table_with_aamr()
# Key fix: Use the SAME population-partition vars for deduping Population:
#   - Partition by Sex + (Single Race 6 if present else Race) for OVERALL rates
#   - If grouping by race_unified, partition by Sex only
# ============================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(readr)
})

# ------------------------------------------------------------
# 0) Parsers / helpers
# ------------------------------------------------------------
parse_pop_safe <- function(x) {
  x <- na_if(as.character(x), "Not Applicable")
  readr::parse_number(x, na = c("", "NA", "Not Applicable"))
}

recode_to_std_age <- function(age_code) {
  dplyr::case_when(
    age_code %in% c("85-89", "90-94", "95-99", "100+") ~ "85+",
    TRUE ~ as.character(age_code)
  )
}

fmt_num <- function(x, digits = 2) ifelse(is.na(x), NA_character_,
                                         formatC(x, format = "f", digits = digits, big.mark = ","))
fmt_int <- function(x) ifelse(is.na(x), NA_character_,
                              formatC(round(x), format = "f", digits = 0, big.mark = ","))
fmt_n_pct <- function(n, pct, pct_digits = 1) {
  ifelse(
    is.na(n) | is.na(pct),
    NA_character_,
    paste0(fmt_int(n), " (", formatC(pct, format = "f", digits = pct_digits), "%)")
  )
}

# ------------------------------------------------------------
# 1) 2000 US standard weights (ALL ages)
# (use your std_pop version here; replace if you want std_million split version)
# ------------------------------------------------------------
std2000_allages <- tibble(
    age_code = c(
      "1","1-4",
      "5-9","10-14",
      "15-19","20-24",
      "25-29","30-34",
      "35-39","40-44",
      "45-49","50-54",
      "55-59","60-64",
      "65-69","70-74",
      "75-79","80-84",
      "85+"
    ),
    std_million = c(
      13818, 55317,
      145565/2, 145565/2,
      138646/2, 138646/2,
      135573/2, 135573/2,
      162613/2, 162613/2,
      134834/2, 134834/2,
      87247/2,  87247/2,
      66037/2,  66037/2,
      44842/2,  44842/2,
      15508
    )
  ) %>%
    mutate(
      std_million = round(std_million),
      weight = std_million / sum(std_million)
    )

# age band map for your age4 summary rows
age_band_map <- tibble(
  age_code = std2000_allages$age_code,
  age4 = c(
    rep("<20", 5),
    rep("20_44", 5),
    rep("45_64", 4),
    rep("65_plus", 5)
  )
)

# ------------------------------------------------------------
# 2) Prep (common cleaning)
# ------------------------------------------------------------
prep_common <- function(df, year_min = 2010, year_max = 2025) {
  df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      YearCode   = suppressWarnings(as.integer(`Year Code`)),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code   = recode_to_std_age(`Five-Year Age Groups Code`),

      race_source = coalesce(`Single Race 6`, Race),
      race_unified = case_when(
        race_source == "American Indian or Alaska Native" ~ "American Indian or Alaska Native",
        race_source %in% c("Asian", "Asian or Pacific Islander") ~ "Asian",
        race_source == "Native Hawaiian or Other Pacific Islander" ~ "Native Hawaiian or Other Pacific Islander",
        race_source == "Black or African American" ~ "Black or African American",
        race_source == "White" ~ "White",
        race_source == "More than one race" ~ "More than one race",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(YearCode >= year_min, YearCode <= year_max) %>%
    filter(age_code %in% std2000_allages$age_code)
}

# ------------------------------------------------------------
# 3) Population partition logic (MATCH make_year_table_with_aamr)
# ------------------------------------------------------------
race_part_var <- function(dat) {
  if ("Single Race 6" %in% names(dat)) return("Single Race 6")
  if ("Race" %in% names(dat))          return("Race")
  return(NULL)
}

# Collapse Year×Age×(group_vars)×(partition vars)
# - Overall / sex-only: partition by Sex + raw race column (Single Race 6 preferred)
# - When grouping by race_unified: partition by Sex only
collapse_for_rates <- function(dat, group_vars = character()) {

  rp <- race_part_var(dat)

  pv <- c("Sex", rp)
  pv <- pv[!is.null(pv)]
  pv <- pv[pv %in% names(dat)]

  if ("race_unified" %in% group_vars) {
    pv <- intersect(pv, "Sex")
  }

  pv <- setdiff(pv, group_vars)

  dat %>%
    group_by(across(all_of(c(group_vars, pv, "YearCode", "age_code")))) %>%
    summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(pop = ifelse(is.infinite(pop), NA_real_, pop))
}

# ------------------------------------------------------------
# 4) AAMR calculators
# ------------------------------------------------------------
calc_aamr_group <- function(dat, group_vars = character()) {

  collapsed <- collapse_for_rates(dat, group_vars)

  # sum across partition vars inside each Year×Age (and group)
  year_age_totals <- collapsed %>%
    group_by(across(all_of(c(group_vars, "YearCode", "age_code")))) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  # pool across years within age_code
  age_totals <- year_age_totals %>%
    group_by(across(all_of(c(group_vars, "age_code")))) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  age_totals %>%
    left_join(std2000_allages, by = "age_code") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(AAMR = sum(age_rate * weight, na.rm = TRUE), .groups = "drop")
}

calc_aamr_ageband <- function(dat) {

  dat2 <- dat %>% left_join(age_band_map, by = "age_code")

  collapsed <- collapse_for_rates(dat2, group_vars = c("age4"))

  year_age_band_totals <- collapsed %>%
    group_by(age4, YearCode, age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  age_band_totals <- year_age_band_totals %>%
    group_by(age4, age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  age_band_totals %>%
    left_join(std2000_allages, by = "age_code") %>%
    group_by(age4) %>%
    mutate(
      w_band  = weight / sum(weight),
      age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)
    ) %>%
    summarise(AAMR = sum(age_rate * w_band, na.rm = TRUE), .groups = "drop")
}

# crude totals (safe; no double-count)
calc_total_deaths_crude <- function(dat) {

  collapsed <- collapse_for_rates(dat, group_vars = character())

  year_age_totals <- collapsed %>%
    group_by(YearCode, age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  totals <- year_age_totals %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  totals %>%
    mutate(crude = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_))
}

# sex deaths pct (use Deaths only; population irrelevant)
calc_sex_deaths_pct <- function(dat) {

  sex_map <- c(
    "Male" = "Male", "Men" = "Male", "M" = "Male",
    "Female" = "Female", "Women" = "Female", "F" = "Female"
  )

  tmp <- dat %>%
    filter(!is.na(Sex)) %>%
    mutate(Sex = recode(Sex, !!!sex_map, .default = Sex)) %>%
    filter(Sex %in% c("Male", "Female"))

  sex_deaths <- tmp %>%
    group_by(Sex) %>%
    summarise(deaths = sum(Deaths, na.rm = TRUE), .groups = "drop")

  male_deaths   <- sex_deaths$deaths[sex_deaths$Sex == "Male"]
  female_deaths <- sex_deaths$deaths[sex_deaths$Sex == "Female"]

  male_deaths   <- ifelse(length(male_deaths)   == 0, 0, male_deaths)
  female_deaths <- ifelse(length(female_deaths) == 0, 0, female_deaths)

  total <- male_deaths + female_deaths

  tibble(
    male_deaths = male_deaths,
    male_pct    = ifelse(total > 0, male_deaths / total * 100, NA_real_),
    female_deaths = female_deaths,
    female_pct    = ifelse(total > 0, female_deaths / total * 100, NA_real_)
  )
}

# ------------------------------------------------------------
# 5) One dataset summary (used by summary_tbl)
# ------------------------------------------------------------
summarise_dataset <- function(df, year_min = 2010, year_max = 2025) {

  dat <- prep_common(df, year_min, year_max)

  sex_deaths_pct <- calc_sex_deaths_pct(dat)

  tot <- calc_total_deaths_crude(dat)

  # OVERALL AAMR (all ages) — this is what must align with make_year_table_with_aamr Total AAMR
  overall_aamr <- calc_aamr_group(dat) %>% summarise(AAMR = first(AAMR))

  sex_aamr <- calc_aamr_group(dat, "Sex") %>%
    select(Sex, AAMR) %>%
    pivot_wider(names_from = Sex, values_from = AAMR)

  age_aamr <- calc_aamr_ageband(dat) %>%
    select(age4, AAMR) %>%
    pivot_wider(names_from = age4, values_from = AAMR)

  race_aamr <- calc_aamr_group(dat %>% filter(!is.na(race_unified)), "race_unified") %>%
    select(race_unified, AAMR) %>%
    pivot_wider(names_from = race_unified, values_from = AAMR)

  list(
    total_deaths = tot$deaths,
    total_pop    = tot$pop,
    crude_rate   = tot$crude,

    overall_aamr = overall_aamr$AAMR,

    male_deaths     = sex_deaths_pct$male_deaths,
    male_deaths_pct = sex_deaths_pct$male_pct,
    female_deaths     = sex_deaths_pct$female_deaths,
    female_deaths_pct = sex_deaths_pct$female_pct,

    male_aamr    = sex_aamr$Male,
    female_aamr  = sex_aamr$Female,

    lt20_aamr    = age_aamr$`<20`,
    age2044_aamr = age_aamr$`20_44`,
    age4564_aamr = age_aamr$`45_64`,
    age65_aamr   = age_aamr$`65_plus`,

    aian_aamr    = race_aamr$`American Indian or Alaska Native`,
    asian_aamr   = race_aamr$Asian,
    black_aamr   = race_aamr$`Black or African American`,
    nhopi_aamr   = race_aamr$`Native Hawaiian or Other Pacific Islander`,
    white_aamr   = race_aamr$White,
    multi_aamr   = race_aamr$`More than one race`
  )
}

# ------------------------------------------------------------
# 6) Build summary_tbl
# ------------------------------------------------------------
# Replace these with your objects:
#   N17, A40, N17A40_RF (or N17A40)
arf  <- summarise_dataset(N17)
seps <- summarise_dataset(A40)
both <- summarise_dataset(N17A40_RF)

summary_tbl <- tibble(
  Metric = c(
    "Total deaths (standardizable ages), n",
    "Male deaths, n (%)",
    "Female deaths, n (%)",
    "Total population (standardizable ages)",
    "Crude mortality rate (standardizable ages), /100,000",
    "AAMR (all ages), /100,000",
    "Male AAMR (all ages), /100,000",
    "Female AAMR (all ages), /100,000",
    "Age categories",
    "  Age <20 AAMR, /100,000",
    "  Age 20–44 AAMR, /100,000",
    "  Age 45–64 AAMR, /100,000",
    "  Age ≥65 AAMR, /100,000",
    "Race and ethnicity",
    "  American Indian or Alaska Native AAMR, /100,000",
    "  Asian AAMR, /100,000",
    "  Black or African American AAMR, /100,000",
    "  Native Hawaiian or Other Pacific Islander AAMR, /100,000",
    "  White AAMR, /100,000",
    "  More than one race AAMR, /100,000"
  ),

  AKI = c(
    fmt_int(arf$total_deaths),
    fmt_n_pct(arf$male_deaths,   arf$male_deaths_pct),
    fmt_n_pct(arf$female_deaths, arf$female_deaths_pct),
    fmt_int(arf$total_pop),
    fmt_num(arf$crude_rate),
    fmt_num(arf$overall_aamr),
    fmt_num(arf$male_aamr),
    fmt_num(arf$female_aamr),
    NA,
    fmt_num(arf$lt20_aamr),
    fmt_num(arf$age2044_aamr),
    fmt_num(arf$age4564_aamr),
    fmt_num(arf$age65_aamr),
    NA,
    fmt_num(arf$aian_aamr),
    fmt_num(arf$asian_aamr),
    fmt_num(arf$black_aamr),
    fmt_num(arf$nhopi_aamr),
    fmt_num(arf$white_aamr),
    fmt_num(arf$multi_aamr)
  ),

  Sepsis = c(
    fmt_int(seps$total_deaths),
    fmt_n_pct(seps$male_deaths,   seps$male_deaths_pct),
    fmt_n_pct(seps$female_deaths, seps$female_deaths_pct),
    fmt_int(seps$total_pop),
    fmt_num(seps$crude_rate),
    fmt_num(seps$overall_aamr),
    fmt_num(seps$male_aamr),
    fmt_num(seps$female_aamr),
    NA,
    fmt_num(seps$lt20_aamr),
    fmt_num(seps$age2044_aamr),
    fmt_num(seps$age4564_aamr),
    fmt_num(seps$age65_aamr),
    NA,
    fmt_num(seps$aian_aamr),
    fmt_num(seps$asian_aamr),
    fmt_num(seps$black_aamr),
    fmt_num(seps$nhopi_aamr),
    fmt_num(seps$white_aamr),
    fmt_num(seps$multi_aamr)
  ),

  `AKI & sepsis` = c(
    fmt_int(both$total_deaths),
    fmt_n_pct(both$male_deaths,   both$male_deaths_pct),
    fmt_n_pct(both$female_deaths, both$female_deaths_pct),
    fmt_int(both$total_pop),
    fmt_num(both$crude_rate),
    fmt_num(both$overall_aamr),
    fmt_num(both$male_aamr),
    fmt_num(both$female_aamr),
    NA,
    fmt_num(both$lt20_aamr),
    fmt_num(both$age2044_aamr),
    fmt_num(both$age4564_aamr),
    fmt_num(both$age65_aamr),
    NA,
    fmt_num(both$aian_aamr),
    fmt_num(both$asian_aamr),
    fmt_num(both$black_aamr),
    fmt_num(both$nhopi_aamr),
    fmt_num(both$white_aamr),
    fmt_num(both$multi_aamr)
  )
)

summary_tbl

```

```{r}
get_summary_total <- function(summary_tbl, outcome_col,
                              metric_name = "Total deaths (standardizable ages), n") {
  summary_tbl %>%
    filter(Metric == metric_name) %>%
    pull(!!sym(outcome_col)) %>%
    readr::parse_number()
}

# ---- 1) totals from summary_tbl (these are the denominators) ----
summary_totals <- tibble(
  dataset = c("AKI", "Sepsis", "AKI & Sepsis"),
  total_deaths = c(
    get_summary_total(summary_tbl, "AKI"),
    get_summary_total(summary_tbl, "Sepsis"),
    get_summary_total(summary_tbl, "AKI & sepsis")   # <-- your column name
  )
)

# ---- 2) medical deaths from pod (sum across periods) ----
pod_med <- pod %>%
  mutate(
    place_group = case_when(
      str_detect(place_group, "^Medical")    ~ "Medical Facility",
      str_detect(place_group, "^Non-medical")~ "Non-medical facility",
      TRUE ~ place_group
    )
  ) %>%
  filter(place_group == "Medical Facility") %>%
  group_by(dataset) %>%
  summarise(medical_deaths = sum(total_deaths, na.rm = TRUE), .groups = "drop")

# ---- 3) compute non-medical = total - medical; format n (x.x%) ----
pod_overall <- summary_totals %>%
  left_join(pod_med, by = "dataset") %>%
  mutate(
    medical_deaths = coalesce(medical_deaths, 0),
    nonmed_deaths  = pmax(total_deaths - medical_deaths, 0),

    med_pct    = ifelse(total_deaths > 0, 100 * medical_deaths / total_deaths, NA_real_),
    nonmed_pct = ifelse(total_deaths > 0, 100 * nonmed_deaths  / total_deaths, NA_real_),

    `  Medical Facility, n (%)` = paste0(
      formatC(round(medical_deaths, 0), format = "f", digits = 0, big.mark = ","),
      " (", formatC(med_pct, format = "f", digits = 1), "%)"
    ),
    `  Non-medical facility, n (%)` = paste0(
      formatC(round(nonmed_deaths, 0), format = "f", digits = 0, big.mark = ","),
      " (", formatC(nonmed_pct, format = "f", digits = 1), "%)"
    )
  ) %>%
  select(dataset, `  Medical Facility, n (%)`, `  Non-medical facility, n (%)`) %>%
  pivot_longer(
    cols = starts_with("  "),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  pivot_wider(names_from = dataset, values_from = Value) %>%
  rename(`AKI & sepsis` = `AKI & Sepsis`)  # match your summary_tbl column name

# ---- 4) add header row + append to summary_tbl ----
pod_header <- tibble(
  Metric = "Place of death",
  AKI = NA_character_,
  Sepsis = NA_character_,
  `AKI & sepsis` = NA_character_
)

pod_section <- bind_rows(pod_header, pod_overall)

summary_tbl2 <- bind_rows(summary_tbl, pod_section)

summary_tbl2
```

## Export

```{r}
library(openxlsx)

wb <- createWorkbook()

addWorksheet(wb, "AKI_year")
writeData(wb, "AKI_year", table_N17_aamr)

addWorksheet(wb, "sepsis_year")
writeData(wb, "sepsis_year", table_A40_aamr)

addWorksheet(wb, "AKI_sepsis_year")
writeData(wb, "AKI_sepsis_year", table_N17A40_renal_aamr)

addWorksheet(wb, "AKI_period_summary")
writeData(wb, "AKI_period_summary", arf_period$publication_table)

addWorksheet(wb, "Sepsis_period_summary")
writeData(wb, "Sepsis_period_summary", sepsis_period$publication_table)

addWorksheet(wb, "AKI_Sepsis_period_summary")
writeData(wb, "AKI_Sepsis_period_summary", renal_only_from_combined$publication_table)

addWorksheet(wb, "Deaths_by_age")
writeData(wb, "Deaths_by_age", deaths_table_age)

addWorksheet(wb, "Rate_by_age")
writeData(wb, "Rate_by_age", rate_table_age)

addWorksheet(wb, "Deaths_by_sex")
writeData(wb, "Deaths_by_sex", deaths_table_sex)

addWorksheet(wb, "Crude_rate_by_sex")
writeData(wb, "Crude_rate_by_sex", aamr_table_sex)

addWorksheet(wb, "ARF_Sepsis_overlap_rates")
writeData(wb, "ARF_Sepsis_overlap_rates", overlap_table_aamr)

addWorksheet(wb, "Overall_Summary")
writeData(wb, "Overall_Summary", summary_tbl)

saveWorkbook(
  wb,
  file = "Supplementary_Tables_ARF_Sepsis_US_2010_2025.xlsx",
  overwrite = TRUE
)

```

```{r}
df_to_flex <- function(df, font_size = 9) {
  flextable(df) |>
    theme_booktabs() |>
    fontsize(size = font_size, part = "all") |>
    autofit() |>
    set_table_properties(layout = "autofit")
}

pick_font <- function(df) {
  p <- ncol(df)
  if (p >= 14) 7 else if (p >= 11) 8 else 9
}

```

```{r}
add_table <- function(doc, title, df) {
  doc |>
    body_add_par(title, style = "heading 1") |>
    body_add_flextable(
      df_to_flex(df, font_size = pick_font(df))
    ) |>
    body_add_par("")
}

```

```{r}
library(officer)
library(flextable)

# ---- flextable helper (robust across versions) ----
df_to_flex <- function(df, font_size = 10) {
  flextable(df) |>
    theme_booktabs() |>
    fontsize(size = font_size, part = "all") |>
    autofit() |>
    set_table_properties(layout = "autofit")
}

# Optional: auto-shrink very wide tables a bit
pick_font <- function(df) {
  p <- ncol(df)
  if (p >= 12) 8 else if (p >= 9) 9 else 10
}

add_table_section <- function(doc, title, df) {
  doc <- body_add_par(doc, title, style = "heading 1")
  doc <- body_add_flextable(doc, df_to_flex(df, font_size = pick_font(df)))
  doc <- body_add_par(doc, "", style = "Normal") # spacer
  doc
}

doc <- read_docx()

# -------------------------
# Year tables
# -------------------------
doc <- add_table_section(doc, "AKI_year", table_N17_aamr)
doc <- add_table_section(doc, "sepsis_year", table_A40_aamr)
doc <- add_table_section(doc, "AKI_sepsis_year", table_N17A40_renal_aamr)

# -------------------------
# Period summaries
# -------------------------
doc <- add_table_section(doc, "AKI_period_summary", arf_period$publication_table)
doc <- add_table_section(doc, "Sepsis_period_summary", sepsis_period$publication_table)
doc <- add_table_section(doc, "AKI_Sepsis_period_summary",renal_only_from_combined$publication_table)

# -------------------------
# Age / Sex breakdowns
# -------------------------
doc <- add_table_section(doc, "Deaths_by_age", deaths_table_age)
doc <- add_table_section(doc, "AAMR_by_age", rate_table_age)        # <-- change name if your object differs
doc <- add_table_section(doc, "Deaths_by_sex", deaths_table_sex)
doc <- add_table_section(doc, "AAMR_by_sex", aamr_table_sex)

# -------------------------
# Overlap + Overall summary
# -------------------------
doc <- add_table_section(doc, "AKI_Sepsis_overlap_rates", overlap_table_aamr_allages)
doc <- add_table_section(doc, "Overall_Summary", summary_tbl2)

# -------------------------
# Save
# -------------------------
print(doc, target = "Supplementary_Tables_AKI_Sepsis_US_2010_2025.docx")


```

## Trend

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(readr)
})

# ============================================================
# 0) helpers
# ============================================================
parse_pop_safe <- function(x) {
  x <- na_if(as.character(x), "Not Applicable")
  readr::parse_number(x, na = c("", "NA", "Not Applicable"))
}

recode_to_std_age <- function(age_code) {
  age_code <- as.character(age_code)
  dplyr::case_when(
    age_code %in% c("85-89", "90-94", "95-99", "100+") ~ "85+",
    TRUE ~ age_code
  )
}

# ============================================================
# 1) 2000 US standard weights (ALL ages; your "split bins" version)
# ============================================================
std2000_allages <- tibble(
  age_code = c(
    "1","1-4",
    "5-9","10-14",
    "15-19","20-24",
    "25-29","30-34",
    "35-39","40-44",
    "45-49","50-54",
    "55-59","60-64",
    "65-69","70-74",
    "75-79","80-84",
    "85+"
  ),
  std_million = c(
    13818, 55317,
    145565/2, 145565/2,
    138646/2, 138646/2,
    135573/2, 135573/2,
    162613/2, 162613/2,
    134834/2, 134834/2,
    87247/2,  87247/2,
    66037/2,  66037/2,
    44842/2,  44842/2,
    15508
  )
) %>%
  mutate(
    std_million = round(std_million),
    weight = std_million / sum(std_million)
  )

# ---- map 5-year age_code -> 4 bands (age4)
age_band_map <- tibble(
  age_code = std2000_allages$age_code,
  age4 = c(
    rep("<20",     5),
    rep("20_44",   5),
    rep("45_64",   4),
    rep("65_plus", 5)
  )
)

# ============================================================
# 2) identify partition vars causing duplicated population
#    (match make_year_table_with_aamr logic)
# ============================================================
get_part_vars <- function(dat) {
  pv <- intersect(c("Sex", "Single Race 6", "Race"), names(dat))
  if (all(c("Race", "Single Race 6") %in% pv)) pv <- setdiff(pv, "Race")
  pv
}

# ============================================================
# 3) canonical prep
# ============================================================
prep_base <- function(df, year_min = 2010, year_max = 2025) {
  df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      YearCode   = suppressWarnings(as.integer(`Year Code`)),
      YearKey    = as.character(YearCode),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code   = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    filter(!is.na(YearCode), YearCode >= year_min, YearCode <= year_max) %>%
    filter(age_code %in% std2000_allages$age_code)
}

# ============================================================
# 4) canonical "YA" table: YearCode × age_code with DEDUPED POP
#    This is the one table BOTH pipelines must use.
# ============================================================
build_ya_dedup <- function(base_dat) {
  pv <- get_part_vars(base_dat)

  base_dat %>%
    group_by(across(all_of(c("YearCode", "age_code", pv)))) %>%
    summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(pop = ifelse(is.infinite(pop), NA_real_, pop)) %>%
    group_by(YearCode, age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )
}

# ============================================================
# 5) AAMR (ALL AGES) from YA
# ============================================================
calc_aamr_allages_by_year <- function(ya) {
  ya %>%
    left_join(std2000_allages, by = "age_code") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    group_by(YearCode) %>%
    summarise(AAMR_allages = sum(age_rate * weight, na.rm = TRUE), .groups = "drop") %>%
    mutate(YearKey = as.character(YearCode)) %>%
    select(YearKey, AAMR_allages)
}

calc_aamr_allages_total <- function(ya) {
  ya %>%
    group_by(age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(std2000_allages, by = "age_code") %>%
    mutate(age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    summarise(
      YearKey = "Total",
      AAMR_allages = sum(age_rate * weight, na.rm = TRUE)
    )
}

# ============================================================
# 6) AAMR (BY AGE BAND) from YA
#    - weights are renormalized within each band
# ============================================================
calc_aamr_band_by_year <- function(ya) {
  ya %>%
    left_join(age_band_map, by = "age_code") %>%
    left_join(std2000_allages %>% select(age_code, weight), by = "age_code") %>%
    group_by(YearCode, age4) %>%
    mutate(
      w_band   = weight / sum(weight),
      age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)
    ) %>%
    summarise(AAMR_band = sum(age_rate * w_band, na.rm = TRUE), .groups = "drop") %>%
    mutate(YearKey = as.character(YearCode)) %>%
    select(YearKey, age4, AAMR_band)
}

calc_aamr_band_total <- function(ya) {
  ya %>%
    group_by(age_code) %>%
    summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(age_band_map, by = "age_code") %>%
    left_join(std2000_allages %>% select(age_code, weight), by = "age_code") %>%
    group_by(age4) %>%
    mutate(
      w_band   = weight / sum(weight),
      age_rate = ifelse(pop > 0, (deaths / pop) * 1e5, NA_real_)
    ) %>%
    summarise(
      YearKey = "Total",
      AAMR_band = sum(age_rate * w_band, na.rm = TRUE),
      .groups = "drop"
    )
}

# ============================================================
# 7) aligned wrapper: returns BOTH overall-AAMR and band-AAMR tables
# ============================================================
aamr_allages_and_bands <- function(df, year_min = 2010, year_max = 2025) {

  base <- prep_base(df, year_min, year_max)
  ya   <- build_ya_dedup(base)

  # overall AAMR (year + total)
  allages_year  <- calc_aamr_allages_by_year(ya) %>%
    complete(YearKey = as.character(year_min:year_max), fill = list(AAMR_allages = NA_real_))
  allages_total <- calc_aamr_allages_total(ya)
  allages_tbl   <- bind_rows(allages_year, allages_total) %>%
    mutate(YearKey = factor(YearKey, levels = c(as.character(year_min:year_max), "Total"))) %>%
    arrange(YearKey)

  # band AAMR (year + total)
  band_year  <- calc_aamr_band_by_year(ya) %>%
    complete(
      YearKey = as.character(year_min:year_max),
      age4    = c("<20","20_44","45_64","65_plus"),
      fill    = list(AAMR_band = NA_real_)
    )
  band_total <- calc_aamr_band_total(ya)
  band_tbl   <- bind_rows(band_year, band_total) %>%
    mutate(
      YearKey = factor(YearKey, levels = c(as.character(year_min:year_max), "Total")),
      age4    = factor(age4, levels = c("<20","20_44","45_64","65_plus"))
    ) %>%
    arrange(YearKey, age4)

  list(
    ya = ya,                 # the canonical deduped denominators
    allages = allages_tbl,   # YearKey + Total overall AAMR
    bands   = band_tbl       # YearKey + Total band AAMR
  )
}

# ============================================================
# 8) Use it to build the SAME outputs you had before
#    (example for your 3 datasets)
# ============================================================
out_aki   <- aamr_allages_and_bands(N17,    2010, 2025)
out_sep   <- aamr_allages_and_bands(A40,    2010, 2025)
out_both  <- aamr_allages_and_bands(N17A40_RF, 2010, 2025)

# ---- overall AAMR by year (wide), aligned across datasets ----
overall_aamr_tbl <- out_aki$allages %>%
  rename(AKI = AAMR_allages) %>%
  left_join(out_sep$allages %>% rename(Sepsis = AAMR_allages), by = "YearKey") %>%
  left_join(out_both$allages %>% rename(`AKI/sepsis` = AAMR_allages), by = "YearKey") %>%
  mutate(
    `Sepsis among AKI ((3)/(1))` = ifelse(AKI > 0, `AKI/sepsis` / AKI, NA_real_),
    `AKI among Sepsis ((3)/(2))` = ifelse(Sepsis > 0, `AKI/sepsis` / Sepsis, NA_real_)
  )

overall_aamr_tbl

# ---- band AAMR trend in your original "aamr_trend long" style ----
aamr_trend <- bind_rows(
  out_aki$bands  %>% mutate(dataset = "AKI"),
  out_sep$bands  %>% mutate(dataset = "Sepsis"),
  out_both$bands %>% mutate(dataset = "AKI/sepsis")
) %>%
  mutate(
    Year = as.integer(as.character(YearKey)),
    age_group = recode(as.character(age4),
      "<20" = "<20",
      "20_44" = "20–44",
      "45_64" = "45–64",
      "65_plus" = "65+"
    ),
    age_group = factor(age_group, levels = c("<20","20–44","45–64","65+"))
  ) %>%
  filter(!is.na(Year)) %>%   # drops Total for plotting; keep if you want it
  transmute(
    Year,
    dataset,
    age_group,
    AAMR = AAMR_band
  )

aamr_trend

```

```{r}
aamr_trend_plot <- aamr_trend %>%
  filter(age_group %in% c("20–44", "45–64", "65+")) %>%
  mutate(
    age_group = factor(age_group, levels = c("20–44", "45–64", "65+"))
  )
```

```{r}
fig_aamr_color <- ggplot(
  aamr_trend_plot,
  aes(
    x = Year,
    y = AAMR,
    color = age_group
  )
) +
  annotate(
      "rect",
      xmin = 2019.5, xmax = 2022.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey70", alpha = 0.18
    ) +
    annotate(
      "rect",
      xmin = 2022.5, xmax = 2023.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey40", alpha = 0.14
    ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  facet_wrap(~ dataset, nrow = 1, scales = "free_y") +
  scale_color_manual(
    values = c(
    "20–44" = "#1b9e77",
    "45–64" = "#d95f02",
    "65+"   = "#7570b3"
  )
) +
  scale_x_continuous(breaks = seq(2010, 2025, 3)) +
  labs(
    x = "Year",
    y = "AAMR(per 100,000)",
    color = "Age group"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    panel.spacing = unit(1.2, "lines")
  )

fig_aamr_color

```

```{r}
plot_single_dataset <- function(df, disease_label) {
  df1 <- df %>%
    mutate(
      dataset = trimws(dataset),
      age_group = trimws(age_group),
      age_group = recode(
        age_group,
        "20_44" = "20–44",
        "45_64" = "45–64",
        "65_plus" = "65+",
        "lt20" = "<20",
        "<20" = "<20",
        "20–44" = "20–44",
        "45–64" = "45–64",
        "65+" = "65+"
      ),
      age_group = factor(age_group, levels = c("<20", "20–44", "45–64", "65+"))
    ) %>%
    filter(dataset == trimws(disease_label)) %>%
    filter(age_group != "<20") %>%  # ✅ remove <20
    droplevels()

  ggplot(
    df1,
    aes(x = Year, y = AAMR, color = age_group)
  ) +
    annotate(
      "rect",
      xmin = 2019.5, xmax = 2022.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey70", alpha = 0.18
    ) +
    annotate(
      "rect",
      xmin = 2022.5, xmax = 2023.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey40", alpha = 0.14
    ) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 2.2) +
    scale_color_manual(
      values = c(
        "20–44" = "#1b9e77",
        "45–64" = "#d95f02",
        "65+"   = "#7570b3"
      ),
      drop = FALSE
    ) +
    scale_x_continuous(breaks = seq(2010, 2025, 3)) +
    labs(
      x = "Year",
      y = "AAMR(per 100,000)",
      color = "Age group",
      caption = "Grey shading denotes the COVID-19 pandemic period (2020–2022); darker shading denotes the transition year (2023)."
    ) +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold")
    )
}


```

```{r}
p_arf <- plot_single_dataset(
  aamr_trend,
  "AKI"
)

p_sepsis <- plot_single_dataset(
  aamr_trend,
  "Sepsis "
)

p_combo <- plot_single_dataset(
  aamr_trend,
  "AKI/sepsis"
)

p_arf
p_sepsis
p_combo

```

```{r}
ggsave(
  "figures/ Figure_AAMR_Arf.png",
  p_arf,
  width = 10,
  height = 5,
  dpi = 600
)

ggsave(
  "figures/ Figure_AAMR_sepsis.png",
  p_sepsis,
  width = 10,
  height = 5,
  dpi = 600
)

ggsave(
  "figures/ Figure_AAMR_combo.png",
  p_combo,
  width = 10,
  height = 5,
  dpi = 600
)


ggsave(
  "figures/ Figure_AAMR_all.png",
  fig_aamr_color,
  width = 10,
  height = 5,
  dpi = 600
)

```

### Sex

```{r}
calc_sex_aamr_trend <- function(df, dataset_name,
                                year_min = 2010, year_max = 2025) {

  # --- prep like your other pipeline ---
  base <- df %>%
    dplyr::filter(is.na(Notes) | Notes != "Total") %>%
    dplyr::mutate(
      Year     = suppressWarnings(as.integer(`Year Code`)),
      YearKey  = as.character(Year),
      Deaths   = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    dplyr::filter(!is.na(Year), Year >= year_min, Year <= year_max) %>%
    dplyr::filter(age_code %in% std2000_allages$age_code) %>%
    dplyr::filter(!is.na(Sex))

  # --- partition vars: align with your collapse_for_rates logic ---
  rp <- NULL
  if ("Single Race 6" %in% names(base)) rp <- "Single Race 6"
  if ("Race" %in% names(base) && is.null(rp)) rp <- "Race"

  pv <- c("Sex", rp)
  pv <- pv[!is.null(pv)]
  pv <- pv[pv %in% names(base)]

  # --- dedupe Population within strata, then sum to true denominator ---
  ya_sex <- base %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c("YearKey", "Year", "age_code", pv)))) %>%
    dplyr::summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::mutate(pop = dplyr::if_else(is.infinite(pop), NA_real_, pop)) %>%
    dplyr::group_by(Year, YearKey, Sex, age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  # --- AAMR standardization ---
  ya_sex %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::mutate(age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    dplyr::group_by(Year, Sex) %>%
    dplyr::summarise(
      AAMR = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      dataset = trimws(dataset_name),
      Sex = factor(trimws(Sex), levels = c("Male", "Female"))
    ) %>%
    dplyr::arrange(dataset, Sex, Year)
}

aamr_sex_trend <- dplyr::bind_rows(
  calc_sex_aamr_trend(N17,       "AKI"),
  calc_sex_aamr_trend(A40,       "Sepsis"),
  calc_sex_aamr_trend(N17A40_RF, "AKI/sepsis")
)

# ============================================================
# UPDATED plot: trims dataset label to avoid "Sepsis " issues
# ============================================================
plot_sex_trend_single <- function(df, disease_label) {

  df1 <- df %>%
    mutate(dataset = trimws(dataset)) %>%
    filter(dataset == trimws(disease_label))

  ggplot(
    df1,
    aes(x = Year, y = AAMR, color = Sex)
  ) +
    annotate(
      "rect",
      xmin = 2019.5, xmax = 2022.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey70", alpha = 0.18
    ) +
    annotate(
      "rect",
      xmin = 2022.5, xmax = 2023.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey40", alpha = 0.14
    ) +
    geom_line(linewidth = 1.1) +
    geom_point(size = 2.3) +
    scale_color_manual(
      values = c(
        "Male"   = "#0072B2",
        "Female" = "#D55E00"
      ),
      drop = FALSE
    ) +
    scale_x_continuous(breaks = seq(2010, 2025, 3)) +
    labs(
      x = "Year",
      y = "Age-adjusted mortality rate (per 100,000)",
      color = "Sex",
      caption = "Grey shading denotes the COVID-19 pandemic period (2020–2022); darker shading denotes the transition year (2023)."
    ) +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold")
    )
}

```

```{r}
p_sex_arf <- plot_sex_trend_single(
  aamr_sex_trend,
  "AKI"
)

p_sex_sepsis <- plot_sex_trend_single(
  aamr_sex_trend,
  "Sepsis"
)

p_sex_combo <- plot_sex_trend_single(
  aamr_sex_trend,
  "AKI/sepsis"
)

p_sex_arf
p_sex_sepsis
p_sex_combo

```

```{r}
sex_all <- ggplot(
  aamr_sex_trend,
  aes(x = Year, y = AAMR, color = Sex)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ dataset, nrow = 1, scales = "free_y") +
  scale_color_manual(
    values = c("Male" = "#0072B2", "Female" = "#D55E00")
  ) +
  scale_x_continuous(breaks = seq(2010, 2025, 3)) +
    annotate(
      "rect",
      xmin = 2019.5, xmax = 2022.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey70", alpha = 0.18
    ) +
    annotate(
      "rect",
      xmin = 2022.5, xmax = 2023.5,
      ymin = -Inf, ymax = Inf,
      fill = "grey40", alpha = 0.14
    ) +
  labs(
    x = "Year",
    y = "AAMR",
    color = "Sex",
    caption = "Grey shading denotes the COVID-19 pandemic period (2020–2022); darker shading denotes the transition year (2023)."
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
sex_all
```

```{r}
ggsave(
  "figures/ Figure_AAMR_Arf_sex.png",
  p_sex_arf,
  width = 10,
  height = 5,
  dpi = 600
)

ggsave(
  "figures/ Figure_AAMR_sepsis_sex.png",
  p_sex_sepsis,
  width = 10,
  height = 5,
  dpi = 600
)

ggsave(
  "figures/ Figure_AAMR_combo_sex.png",
  p_sex_combo,
  width = 10,
  height = 5,
  dpi = 600
)


ggsave(
  "figures/ Figure_AAMR_all_sex.png",
  sex_all,
  width = 10,
  height = 5,
  dpi = 600
)
```

### Overlap

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

period_map <- tibble(
  period = c("Pre-pandemic (2010–2019)",
             "Pandemic (2020–2022)",
             "Transition (2023)",
             "Post-pandemic (2024–2025)"),
  year_start = c(2010, 2020, 2023, 2024),
  year_end   = c(2019, 2022, 2023, 2025)
)

add_period <- function(df) {
  df %>%
    left_join(period_map, join_by(Year >= year_start, Year <= year_end)) %>%
    filter(!is.na(period)) %>%
    mutate(period = factor(period, levels = period_map$period))
}

parse_pop_safe <- function(x) {
  x <- na_if(x, "Not Applicable")
  readr::parse_number(x, na = c("", "NA", "Not Applicable"))
}

```

```{r}
deaths_by_period <- function(df, label, year_min = 2010, year_max = 2025) {
  df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      Year = as.integer(`Year Code`),
      Deaths = as.numeric(Deaths)
    ) %>%
    filter(Year >= year_min, Year <= year_max) %>%
    group_by(Year) %>%
    summarise(deaths = sum(Deaths, na.rm = TRUE), .groups = "drop") %>%
    add_period() %>%
    group_by(period) %>%
    summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
    mutate(dataset = label)
}

# ----------------------------
# Build overlap decomposition per period
# ----------------------------
arf_p  <- deaths_by_period(N17,    "AKI")
sep_p  <- deaths_by_period(A40,    "Sepsis")
both_p <- deaths_by_period(N17A40_RF, "AKI/sepsis")   # overlap

overlap_period_pct <- arf_p %>%
  select(period, arf = deaths) %>%
  left_join(sep_p  %>% select(period, sepsis = deaths), by = "period") %>%
  left_join(both_p %>% select(period, both = deaths),   by = "period") %>%
  mutate(
    arf_only    = pmax(arf - both, 0),
    sepsis_only = pmax(sepsis - both, 0),
    total_union = arf_only + sepsis_only + both
  ) %>%
  pivot_longer(
    cols = c(arf_only, sepsis_only, both),
    names_to = "category_raw",
    values_to = "value"
  ) %>%
  mutate(
    category = recode(category_raw,
                      arf_only = "AKI",
                      sepsis_only = "Sepsis",
                      both = "AKI/sepsis"),
    category = factor(category, levels = c("AKI", "AKI/sepsis", "Sepsis")),
    pct = ifelse(total_union > 0, value / total_union, NA_real_),
    pct_label = ifelse(!is.na(pct), paste0(sprintf("%.1f", 100 * pct), "%"), NA_character_)
  )
```

```{r}
library(ggalluvial)
library(scales)
plot_df <- overlap_period_pct %>%
  mutate(
    period   = factor(period, levels = period_map$period),
    category = factor(category, levels = c("AKI", "AKI/sepsis", "Sepsis"))
  )

pal <- c(
  "AKI" = "#E6A36A",
  "AKI/sepsis" = "#8F90C7",
  "Sepsis" = "#6EC6B3"
)

p_overlap_clean <- ggplot(
  plot_df,
  aes(axis1 = period, axis2 = category, y = pct)
) +
  geom_alluvium(aes(fill = category),
                alpha = 0.55, width = 0.12) +
  geom_stratum(
    width = 0.18,
    fill = "grey92",
    color = "grey35",
    linewidth = 0.35
  ) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    size = 3.8,
    color = "grey10"
  ) +
  scale_x_discrete(
    limits = c("Period", "Overlap category"),
    expand = c(0.06, 0.06)
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    breaks = seq(0, 1, 0.25),
    expand = expansion(mult = c(0.03, 0.03))
  ) +
  scale_fill_manual(values = pal) +
  labs(
    x = NULL,
    y = "Share within period",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey88", linewidth = 0.3),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "grey25"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "top",
    legend.text = element_text(size = 11),
    plot.margin = margin(10, 18, 10, 18)
  )

p_overlap_clean
```

```{r}
ggsave(
  "figures/ Figure_overlap.png",
  p_overlap_clean,
  width = 12,
  height = 5,
  dpi = 600
)

```

### \<20

```{r}
N17A40_AKI <- N17A40 %>%
  dplyr::filter(stringr::str_detect(`MCD - ICD Sub-Chapter Code`, "^N17"))
```

```{r}
calc_aamr_lt20_trend <- function(df, dataset_name, year_min = 2010, year_max = 2025) {

  lt20_codes <- c("1","1-4","5-9","10-14","15-19")

  base <- df %>%
    dplyr::filter(is.na(Notes) | Notes != "Total") %>%
    dplyr::mutate(
      Year       = suppressWarnings(as.integer(`Year Code`)),
      YearKey    = as.character(Year),
      Deaths     = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code   = trimws(as.character(`Five-Year Age Groups Code`))
    ) %>%
    dplyr::filter(!is.na(Year), Year >= year_min, Year <= year_max) %>%
    dplyr::filter(age_code %in% lt20_codes)

  # partition vars (align with your other dedupe logic)
  rp <- NULL
  if ("Single Race 6" %in% names(base)) rp <- "Single Race 6"
  if ("Race" %in% names(base) && is.null(rp)) rp <- "Race"

  pv <- c("Sex", rp)
  pv <- pv[!is.null(pv)]
  pv <- pv[pv %in% names(base)]

  # 1) dedupe pop within stratum (Year × age × pv) via max
  # 2) sum strata -> true Year × age denominator
  year_age <- base %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c("YearKey","Year","age_code", pv)))) %>%
    dplyr::summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = dplyr::if_else(all(is.na(Population)), NA_real_,
                              suppressWarnings(max(Population, na.rm = TRUE))),
      .groups = "drop"
    ) %>%
    dplyr::mutate(pop = dplyr::if_else(is.infinite(pop), NA_real_, pop)) %>%
    dplyr::group_by(Year, YearKey, age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  year_age %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::group_by(Year) %>%
    dplyr::mutate(
      w_band   = weight / sum(weight, na.rm = TRUE),   # renormalize within <20
      age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)
    ) %>%
    dplyr::summarise(
      AAMR_lt20 = sum(age_rate * w_band, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(dataset = trimws(dataset_name)) %>%
    dplyr::arrange(dataset, Year)
}

aamr_lt20_trend <- dplyr::bind_rows(
  calc_aamr_lt20_trend(A40, "Sepsis")
)

p_lt20 <- ggplot(aamr_lt20_trend, aes(x = Year, y = AAMR_lt20, color = dataset)) +
  annotate("rect", xmin = 2019.5, xmax = 2022.5, ymin = -Inf, ymax = Inf,
           fill = "grey70", alpha = 0.18) +
  annotate("rect", xmin = 2022.5, xmax = 2023.5, ymin = -Inf, ymax = Inf,
           fill = "grey40", alpha = 0.14) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = seq(2010, 2025, 3)) +
  labs(
    x = "Year",
    y = "AAMR <20 (per 100,000)",
    color = NULL,
    caption = "Grey shading: 2020–2022 (pandemic); darker shading: 2023 (transition)."
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

p_lt20
```


```{r}
N17A40 %>% filter(`Year Code` %in% c(2016, 2023, 2024)) %>%
  count(`MCD - ICD Sub-Chapter`, `MCD - ICD Sub-Chapter Code`, sort = TRUE)
```

```{r}
yrs_check  <- c(2016, 2023, 2024)
lt20_codes <- c("1","1-4","5-9","10-14","15-19")

# Helper: sum <20 deaths + population for selected years
sum_lt20_deaths_pop <- function(df, label) {
  df %>%
    filter(is.na(Notes) | Notes != "Total") %>%
    mutate(
      Year       = as.integer(`Year Code`),
      Deaths     = as.numeric(Deaths),
      Population = parse_pop_safe(Population),
      age_code  = `Five-Year Age Groups Code`
    ) %>%
    filter(
      Year %in% yrs_check,
      age_code %in% lt20_codes
    ) %>%
    group_by(Year) %>%
    summarise(
      Deaths_lt20 = sum(Deaths, na.rm = TRUE),
      Pop_lt20    = sum(Population, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(dataset = label)
}

# ---- AKI vs AKI & Sepsis ----
lt20_deaths_pop <- bind_rows(
  sum_lt20_deaths_pop(N17,   "AKI"),
  sum_lt20_deaths_pop(N17A40,"AKI & Sepsis")
) %>%
  arrange(Year, dataset)

lt20_deaths_pop
```

```{r}
ggsave(
  "figures/ Figure_p_lt20.png",
  p_lt20,
  width = 10,
  height = 5,
  dpi = 600
)
```

### All

```{r}
calc_aamr_trend_all_ages <- function(df, dataset_name, year_min = 2010, year_max = 2025) {

  base <- df %>%
    dplyr::filter(is.na(Notes) | Notes != "Total") %>%
    dplyr::mutate(
      Year     = suppressWarnings(as.integer(`Year Code`)),
      YearKey  = as.character(Year),
      Deaths   = suppressWarnings(as.numeric(Deaths)),
      Population = parse_pop_safe(Population),
      age_code = recode_to_std_age(`Five-Year Age Groups Code`)
    ) %>%
    dplyr::filter(!is.na(Year), Year >= year_min, Year <= year_max) %>%
    dplyr::filter(age_code %in% std2000_allages$age_code)

  # Partition vars (align with your other logic)
  rp <- NULL
  if ("Single Race 6" %in% names(base)) rp <- "Single Race 6"
  if ("Race" %in% names(base) && is.null(rp)) rp <- "Race"

  pv <- c("Sex", rp)                     # include Sex if present; harmless for "all ages"
  pv <- pv[!is.null(pv)]
  pv <- pv[pv %in% names(base)]

  # 1) dedupe pop within stratum (Year × age × pv)
  # 2) sum strata -> true Year × age denominator
  year_age <- base %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c("YearKey", "Year", "age_code", pv)))) %>%
    dplyr::summarise(
      deaths = sum(Deaths, na.rm = TRUE),
      pop    = suppressWarnings(max(Population, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    dplyr::mutate(pop = dplyr::if_else(is.infinite(pop), NA_real_, pop)) %>%
    dplyr::group_by(Year, YearKey, age_code) %>%
    dplyr::summarise(
      deaths = sum(deaths, na.rm = TRUE),
      pop    = sum(pop,    na.rm = TRUE),
      .groups = "drop"
    )

  year_age %>%
    dplyr::left_join(std2000_allages %>% dplyr::select(age_code, weight), by = "age_code") %>%
    dplyr::mutate(age_rate = dplyr::if_else(pop > 0, (deaths / pop) * 1e5, NA_real_)) %>%
    dplyr::group_by(Year) %>%
    dplyr::summarise(
      AAMR = sum(age_rate * weight, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(dataset = trimws(dataset_name)) %>%
    dplyr::arrange(dataset, Year)
}


```

```{r}
aamr_trend_all_ages <- bind_rows(
  calc_aamr_trend_all_ages(N17,    "AKI"),
  calc_aamr_trend_all_ages(A40,    "Sepsis"),
  calc_aamr_trend_all_ages(N17A40_RF, "AKI/sepsis")
)
```

```{r}
p_aamr_all_ages <- ggplot(aamr_trend_all_ages, aes(x = Year, y = AAMR, color = dataset)) +
  annotate(
    "rect",
    xmin = 2019.5, xmax = 2022.5,
    ymin = -Inf, ymax = Inf,
    fill = "grey70", alpha = 0.18
  ) +
  annotate(
    "rect",
    xmin = 2022.5, xmax = 2023.5,
    ymin = -Inf, ymax = Inf,
    fill = "grey40", alpha = 0.14
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.3) +
  scale_color_manual(
    values = c(
      "AKI" = "#0072B2",
      "Sepsis"              = "#D55E00",
      "AKI/sepsis"        = "#009E73"
    )
  ) +
  scale_x_continuous(breaks = seq(2010, 2025, 3)) +
  labs(
    x = "Year",
    y = "AAMR (per 100,000)",
    color = "Disease",
    caption = "Grey shading: 2020–2022 (pandemic); darker shading: 2023 (transition)."
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )

p_aamr_all_ages

```

```{r}
ggsave("figures/ AAMR_trend_overall_shaded.png", p_aamr_all_ages,
       width = 10, height = 5, dpi = 600)
```

### Baseline

```{r}
overlap_table_aamr
```

```{r}
aamr_long <- overlap_table_aamr %>%
  filter(Year != "Total") %>%
  mutate(Year = as.integer(as.character(Year))) %>%
  pivot_longer(
    cols = c(
      `ARF (AAMR 25–85+)`,
      `Sepsis (AAMR 25–85+)`,
      `ARF & Sepsis (AAMR 25–85+)`
    ),
    names_to = "Outcome",
    values_to = "AAMR"
  )
```

```{r}
growth_summary <- aamr_long %>%
  group_by(Outcome) %>%
  summarise(
    baseline_2010_2019 = mean(AAMR[Year >= 2010 & Year <= 2019], na.rm = TRUE),
    peak_aamr          = max(AAMR, na.rm = TRUE),
    peak_year          = Year[which.max(AAMR)],
    aamr_2025          = AAMR[Year == 2025],
    .groups = "drop"
  ) %>%
  mutate(
    growth_peak_vs_baseline =
      (peak_aamr - baseline_2010_2019) / baseline_2010_2019,
    growth_2025_vs_baseline =
      (aamr_2025 - baseline_2010_2019) / baseline_2010_2019
  )

```

```{r}
growth_table_pub <- growth_summary %>%
  transmute(
    Outcome,
    `Baseline AAMR (2010–2019)` = round(baseline_2010_2019, 2),
    `Peak AAMR` = round(peak_aamr, 2),
    `Peak year` = peak_year,
    `AAMR in 2025` = round(aamr_2025, 2),
    `Growth: Peak vs Baseline (%)` =
      round(growth_peak_vs_baseline * 100, 1),
    `Growth: 2025 vs Baseline (%)` =
      round(growth_2025_vs_baseline * 100, 1)
  )

growth_table_pub

```

```{r}
growth_ft <- flextable(growth_table_pub) |>
  autofit() |>
  set_caption(
    "Relative change in mortality rates at pandemic peak and in 2025 compared with the 2010–2019 average"
  ) |>
  align(align = "center", part = "all") |>
  bold(part = "header") |>
  fontsize(size = 11, part = "all") |>
  theme_booktabs()
```

```{r}
doc <- read_docx() |>
  body_add_par(
    "Supplementary Table: Growth of Mortality Rates Relative to Pre-pandemic Baseline",
    style = "heading 1"
  ) |>
  body_add_flextable(growth_ft)

print(doc, target = "Growth_vs_PrePandemic_Baseline.docx")

```

```{r}
library(writexl)
write_xlsx(
  growth_table_pub,
  path = "Growth_vs_PrePandemic_Baseline.xlsx"
)
```
